---
title: "Melbourne's Urban Pulse: Analysing Pedestrian Movement Across Time and Space"
author: "ETC5543 Commercial Analytics"
format:
  html:
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r setup}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
library(forecast)
library(zoo)

OUT_FIG <- "output/figures"
OUT_TAB <- "output/tables"
DATA_PRO <- "data/processed"
DATA_RAW <- "data/raw"
dir.create(OUT_FIG, recursive = TRUE, showWarnings = FALSE)
dir.create(OUT_TAB, recursive = TRUE, showWarnings = FALSE)

df <- read.csv(file.path(DATA_PRO, "pedestrian_counts_processed.csv"),
               stringsAsFactors = FALSE)
df$sensing_date <- as.Date(df$sensing_date)
df$hourday      <- as.integer(df$hourday)
```

# Abstract

This report analyses 4,562,230 hourly pedestrian observations from 134 sensors across Melbourne's CBD, spanning May 2009 to October 2022. Using harmonic regression, ANOVA, STL decomposition, interrupted time series analysis, multiple regression, and ARIMA modelling, we characterise pedestrian rhythms at daily, weekly, and seasonal scales and quantify the disruptive impact of COVID-19. Key findings include a dual-peak daily pattern (08:00 and 17:00, R² = 0.937), a statistically significant day-of-week effect (ANOVA F = 39.84), seasonal peaks in March and November, a COVID-19 shock reducing volumes to 45.8% of the 2019 baseline with only partial recovery to 63% by 2022, and significant weather effects (temperature and wind speed). Thirteen anomalous days were detected, all corresponding to documented events.

---

# 1. Introduction

Understanding how people move through a city has important implications for urban planning, public transport, retail strategy, and public health. Melbourne consistently ranks among the world's most walkable cities, and its CBD is served by a network of automated pedestrian counting sensors that have generated continuous hourly data since 2009. The City of Melbourne Open Data Portal provides free access to this dataset, which covers both indoor and outdoor locations including key pedestrian corridors such as Collins Street, Swanston Street, Flinders Street Station, Melbourne Central, and Princes Bridge. The dataset spans 13 years, encompassing periods of rapid CBD growth, the pedestrianisation of key streets, and the full trajectory of COVID-19 restrictions.

This report addresses five primary research questions: (1) What are the characteristic temporal patterns in Melbourne's pedestrian flows at daily, weekly, and seasonal scales? (2) How has COVID-19 disrupted these patterns, and has recovery occurred? (3) To what extent do weather variables influence pedestrian behaviour? (4) Which days deviate significantly from expected patterns, and can these be explained? (5) Can future pedestrian volumes be reliably forecast?

---

# 2. Data and Methodology

## 2.1 Data Sources

Hourly pedestrian counts were obtained from the City of Melbourne Open Data Portal (data.melbourne.vic.gov.au), dataset "pedestrian-counting-system-monthly-counts-per-hour" (UID: da_8eaok5, CC BY 4.0). The dataset covers 1 May 2009 to 31 October 2022, totalling 4,562,230 observations across 134 sensors. Weather observations were obtained from the Bureau of Meteorology dataset "bom-weather-at-olympic-park" via the same portal. Sensor metadata were obtained from the "pedestrian-counting-system-sensor-locations" dataset. All data were accessed via the OpenDataSoft Explore API v2.0. Historical pedestrian data were downloaded as a CSV zip file from the exports endpoint; weather and sensor data were fetched via paginated records API calls.

## 2.2 Data Processing

The raw pedestrian CSV was processed using chunked reading to handle the 403 MB file efficiently. Each row was parsed and validated; negative pedestrian counts were set to missing. Dates were decomposed into year, month, day, day of week, and week of year. A COVID-19 phase variable was created based on documented Victorian lockdown dates: Normal (pre-March 2020), COVID Stage 1 (March-May 2020), COVID Stage 2 (July-October 2020), Lockdown 8 (August-October 2021), and Endemic (November 2021 onwards). The processed dataset contains 4,562,230 observations spanning 2009-05-01 to 2022-10-31. For weather analysis, pedestrian data were aggregated to daily totals and merged with daily weather summaries, yielding 239 common observation days without COVID restrictions.

## 2.3 Analytical Methods

Seven analytical methods were applied. Harmonic regression using a two-harmonic Fourier series was used to model the daily pedestrian cycle, capturing the morning and evening peaks. One-way ANOVA with the non-parametric Kruskal-Wallis test was used to assess day-of-week differences. Seasonal-Trend decomposition using Loess (STL) was applied to the daily time series from the top-performing sensor to decompose trend, seasonal, and remainder components. Interrupted time series analysis using segmented regression was applied to monthly totals to quantify the COVID-19 impact. Multiple linear regression was used to assess weather effects. ARIMA modelling with automatic order selection was applied to monthly totals for forecasting. Rolling Z-score anomaly detection was applied to daily totals to flag statistically anomalous days.

---

# 3. Results

## 3.1 Annual Volume and Growth

Annual totals grew substantially from 62.7 million in 2009 to a peak of 266.4 million in 2019, representing growth of approximately 325% over the decade. The fastest growth occurred between 2013 and 2015, coinciding with the Melbourne CBD revitalisation program and pedestrianisation of Collins Street. The year 2020 saw a dramatic collapse to 121.9 million (down 54.2% from 2019) as COVID-19 restrictions took effect. Partial recovery to 139.3 million occurred in 2021, and 167.7 million in 2022.

```{r annual-volume}
#| label: annual-volume
#| fig-cap: "Total annual pedestrian counts, 2009 to 2022. The steady growth trend from 2009 to 2019 is followed by a sharp collapse in 2020 and gradual recovery through 2021-2022."
annual <- df %>%
  group_by(year) %>%
  summarise(total = sum(pedestriancount), .groups = "drop")

ggplot(annual, aes(x = year, y = total/1e6, fill = year >= 2020)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1fM", total/1e6)), vjust = -0.5, size = 2.8) +
  scale_fill_manual(values = c("FALSE" = "#00205B", "TRUE" = "#DA291C"), guide = "none") +
  scale_x_continuous(breaks = 2009:2022) +
  labs(x = "Year", y = "Annual pedestrians (millions)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 11)
```

## 3.2 Daily and Weekly Patterns

The hourly pattern exhibits a clear dual-peak structure on weekdays: a morning peak centred around 08:00 and an evening peak around 17:00. The evening peak is broader and slightly higher than the morning peak. The overnight hours (00:00 to 06:00) show near-zero counts, as expected for a CBD location. A harmonic regression model with two Fourier terms was fitted to mean hourly counts, achieving R-squared of 0.937, indicating that the sinusoidal specification captures the daily cycle very well.

```{r hourly-harmonic}
#| label: hourly-harmonic
#| fig-cap: "Mean hourly pedestrian counts with harmonic regression fit (two Fourier pairs). The model achieves R squared of 0.937, capturing the morning peak at 08:00 and evening peak at 17:00."
hourly_avg <- df %>%
  group_by(hourday) %>%
  summarise(mean_count = mean(pedestriancount), .groups = "drop") %>%
  filter(hourday >= 0 & hourday <= 23)

mod_hourly <- lm(mean_count ~
                 sin(2*pi*hourday/24) + cos(2*pi*hourday/24) +
                 sin(4*pi*hourday/24) + cos(4*pi*hourday/24),
                 data = hourly_avg)
hourly_avg$fitted <- predict(mod_hourly, newdata = hourly_avg)

ggplot(hourly_avg, aes(x = hourday)) +
  geom_point(aes(y = mean_count), color = "#00205B", alpha = 0.7, size = 2.5) +
  geom_line(aes(y = fitted), color = "#DA291C", linewidth = 1.5) +
  labs(x = "Hour of day (0-23)", y = "Mean pedestrian count per hour") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  theme_minimal(base_size = 11)

sink(file.path(OUT_TAB, "tab_r1_hourly_harmonic_coef.txt"))
cat("=== Harmonic Regression Coefficients ===\n")
print(summary(mod_hourly)$coefficients)
cat("\nR-squared:", summary(mod_hourly)$r.squared, "\n")
sink()
```

Day-of-week effects are pronounced. One-way ANOVA confirms that mean daily pedestrian counts differ significantly across days of the week (F = 39.84, df = 6, p < 2e-16). The Kruskal-Wallis non-parametric test confirms this result (chi-squared = 244.5, df = 6, p = 6.0e-50). Saturday records the highest mean count, followed by Thursday, while Sunday records the lowest. This pattern reflects the mixed nature of Melbourne's CBD: Saturday brings leisure visitors who compensate for the absence of commuters, while Sunday's reduced trading hours and lack of commuter traffic produces the lowest volumes.

```{r dow-anova}
#| label: dow-anova
#| fig-cap: "Distribution of daily pedestrian totals by day of week. Saturday has the highest median; Sunday has the lowest. The non-parametric Kruskal-Wallis test confirms the day-of-week effect is highly significant (p < 1e-49)."
daily <- df %>%
  group_by(sensing_date, day_of_week) %>%
  summarise(daily_total = sum(pedestriancount), .groups = "drop")

mod_anova <- aov(daily_total ~ factor(day_of_week), data = daily)
kw <- kruskal.test(daily_total ~ factor(day_of_week), data = daily)

dow_names <- c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")
ggplot(daily, aes(x = factor(day_of_week, labels = dow_names),
                  y = daily_total/1e6, fill = day_of_week >= 5)) +
  geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.5) +
  scale_fill_manual(values = c("FALSE" = "#00205B", "TRUE" = "#9B5DE5"), guide = "none") +
  labs(x = "Day of week", y = "Daily pedestrians (millions)") +
  theme_minimal(base_size = 11)

sink(file.path(OUT_TAB, "tab_r2_dow_anova.txt"))
cat("=== Day-of-Week ANOVA ===\n\n")
cat("ANOVA table:\n"); print(summary(mod_anova))
cat("\nKruskal-Wallis chi2:", kw$statistic, "df:", kw$parameter, "p:", kw$p.value, "\n")
sink()
```

## 3.3 Seasonal Patterns

March records the highest seasonal index (114), followed by November (112). June records the lowest (90). The March peak likely reflects the Melbourne Autumn Festival, pleasant autumn weather drawing visitors to the CBD, and foot traffic associated with the start of the academic year. The November peak coincides with the Melbourne Cup Carnival, one of Australia's largest sporting and social events. The June trough reflects winter conditions and the end-of-financial-year period.

```{r seasonality}
#| label: seasonality
#| fig-cap: "Monthly seasonality index. Values above 100 indicate above-average months; values below 100 indicate below-average months. March (index = 114) and November (112) are peak months; June (90) is the trough."
monthly_all <- df %>%
  group_by(year, month) %>%
  summarise(total = sum(pedestriancount), .groups = "drop")
overall_mean <- mean(monthly_all$total)
monthly_all$seas_index <- monthly_all$total / overall_mean * 100
month_seas <- monthly_all %>%
  group_by(month) %>%
  summarise(mean_index = mean(seas_index), sd_index = sd(seas_index), .groups = "drop")

ggplot(month_seas, aes(x = factor(month, labels = month.abb),
                       y = mean_index, fill = mean_index > 100)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey40") +
  geom_text(aes(label = sprintf("%.1f", mean_index)), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("TRUE" = "#00205B", "FALSE" = "#F4A261"), guide = "none") +
  labs(x = "Month", y = "Seasonality index (100 = annual mean)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 11)

sink(file.path(OUT_TAB, "tab_r4_monthly_seasonality.txt"))
cat("=== Monthly Seasonality Index ===\n"); print(month_seas); sink()
write.csv(month_seas, file.path(OUT_TAB, "tab_r4_monthly_seasonality.csv"), row.names = FALSE)
```

## 3.4 Trend, Seasonal, and Remainder Decomposition

STL decomposition was applied to the daily time series from the Melbourne Town Hall (West) sensor, which recorded the highest total counts over the study period. The variance decomposition shows that the trend component accounts for 70.3% of total variation, the seasonal (weekly) component accounts for 6.5%, and the remainder accounts for 29.9%. The large remainder component indicates substantial irregular variation from events, weather, and other unmeasured factors.

```{r stl}
#| label: stl
#| fig-cap: "STL decomposition of daily pedestrian counts from the Melbourne Town Hall (West) sensor. The trend component captures long-run growth, the COVID collapse, and partial recovery. The seasonal component reveals the weekly pattern. The remainder captures irregular fluctuations."
TOP_SENSOR <- df %>%
  group_by(sensor_name) %>%
  summarise(total = sum(pedestriancount)) %>%
  slice_max(total, n = 1) %>%
  pull(sensor_name)

daily_top <- df %>%
  filter(sensor_name == TOP_SENSOR) %>%
  group_by(sensing_date) %>%
  summarise(count = sum(pedestriancount)) %>%
  complete(sensing_date = seq.Date(min(sensing_date), max(sensing_date), by = "day"),
           fill = list(count = 0)) %>%
  arrange(sensing_date)

ts_daily <- ts(daily_top$count, frequency = 7)
stl_fit <- stl(ts_daily, s.window = "periodic", t.window = 365, robust = TRUE)

tot_var       <- var(ts_daily, na.rm = TRUE)
trend_var     <- var(stl_fit$time.series[,"trend"],     na.rm = TRUE)
seasonal_var  <- var(stl_fit$time.series[,"seasonal"],  na.rm = TRUE)
remainder_var <- var(stl_fit$time.series[,"remainder"], na.rm = TRUE)

par(mar = c(3, 3, 2, 1), mgp = c(1.5, 0.5, 0), tck = -0.01)
plot(stl_fit, main = paste("STL Decomposition —", TOP_SENSOR), cex.main = 0.9)

comp_df <- as.data.frame(stl_fit$time.series)
comp_df$date <- daily_top$sensing_date
write.csv(comp_df, file.path(OUT_TAB, "tab_r3_stl_components.csv"), row.names = FALSE)

sink(file.path(OUT_TAB, "tab_r3_stl_summary.txt"))
cat("=== STL Decomposition Summary ===\n")
cat("Trend contribution:", round(trend_var/tot_var*100,1), "%\n")
cat("Seasonal contribution:", round(seasonal_var/tot_var*100,1), "%\n")
cat("Remainder contribution:", round(remainder_var/tot_var*100,1), "%\n")
sink()
```

## 3.5 Top Sensor Analysis

The top 10 sensors by total pedestrian volume were analysed. Melbourne Town Hall (West) consistently records the highest volumes, reflecting its position at the intersection of Collins Street and Swanston Street, two of Melbourne's busiest pedestrian corridors. Collins Street sensors show strong seasonal variation, with higher volumes in warmer months. Indoor sensors such as Melbourne Central show less seasonal amplitude, consistent with climate-controlled environments.

---

# 4. Weather Effects on Pedestrian Volumes

Melbourne is well known for its variable weather. We investigated whether weather variables systematically affect pedestrian volumes by regressing daily totals on mean temperature, total rainfall, mean relative humidity, and mean wind speed.

The regression was estimated on days without COVID restrictions, covering the period where both pedestrian and weather data were available. Temperature and wind speed are statistically significant at the 0.1% level. Each additional degree Celsius is associated with approximately 14,260 additional pedestrians per day. Each additional km/h of mean wind speed is associated with approximately 16,140 fewer pedestrians per day. Rainfall and humidity are not statistically significant, possibly attributable to Melbourne's extensive network of covered walkways and tram shelters.

The overall model R-squared of 0.289 indicates that weather variables explain about 29% of the day-to-day variation in pedestrian volumes.

---

# 5. COVID-19 Impact and Recovery

Melbourne experienced some of the longest COVID-19 lockdowns of any city globally. Stage 3 restrictions began on 31 March 2020, and Stage 4 "hard lockdown" (8pm curfew, 5km travel restriction, closure of non-essential retail) lasted from 2 August to 27 October 2020. A second major lockdown occurred from August to October 2021 during the Delta variant outbreak.

An interrupted time series model was fitted to monthly pedestrian totals from January 2017 to October 2022. The pre-COVID slope of approximately +175,690 pedestrians per month represents strong underlying growth. The COVID level change of approximately -18.1 million pedestrians per month represents the immediate shock of restrictions. The recovery trajectory is dramatically slower than the pre-COVID growth trend, and even by October 2022, monthly volumes had not recovered to pre-COVID levels.

In 2020, volumes collapsed to 45.8% of the 2019 level. Recovery to 52.3% occurred in 2021 and 63.0% in 2022. No year has recovered to the 2019 benchmark, suggesting that structural changes in how Melburnians work, particularly the widespread adoption of hybrid and remote work arrangements, may have permanently altered CBD pedestrian patterns.

```{r covid-its}
#| label: covid-its
#| fig-cap: "Interrupted time series model of monthly pedestrian totals. The pre-COVID upward trend (blue) is interrupted at March 2020. The red fitted line shows the post-COVID trajectory. The shaded bands indicate lockdown periods. Recovery by October 2022 is only 63% of the 2019 baseline."
df_covid <- df[df$year %in% 2017:2023, ]
monthly_ts <- df_covid %>%
  group_by(year, month) %>%
  summarise(total = sum(pedestriancount), .groups = "drop") %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-"))) %>%
  arrange(date) %>%
  mutate(t = seq_along(date))

monthly_ts$post_covid <- as.integer(monthly_ts$date >= "2020-03-01")
monthly_ts$post_time  <- pmax(0, as.numeric(monthly_ts$date - as.Date("2020-03-01")))
its_mod <- lm(total ~ t + post_covid + post_time, data = monthly_ts)
monthly_ts$fitted <- predict(its_mod, newdata = monthly_ts)

ggplot(monthly_ts, aes(x = date)) +
  geom_point(aes(y = total/1e6), color = "#00205B", alpha = 0.7, size = 1.5) +
  geom_line(aes(y = total/1e6), alpha = 0.4, linewidth = 0.4) +
  geom_line(aes(y = fitted/1e6), color = "#DA291C", linewidth = 1.2) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "#DA291C") +
  annotate("rect", xmin = as.Date("2020-03-16"), xmax = as.Date("2020-06-01"),
           ymin = 0, ymax = Inf, alpha = 0.08, fill = "#DA291C") +
  annotate("rect", xmin = as.Date("2021-05-27"), xmax = as.Date("2021-10-21"),
           ymin = 0, ymax = Inf, alpha = 0.08, fill = "#DA291C") +
  labs(x = "Month", y = "Monthly pedestrians (millions)") +
  theme_minimal(base_size = 11)

sink(file.path(OUT_TAB, "tab_r5_covid_its.txt"))
cat("=== Interrupted Time Series ===\n"); print(summary(its_mod)); sink()
```

---

# 6. Anomalous Days

Anomalous days were identified using a rolling 30-day Z-score threshold of |z| > 3 applied to daily totals across all sensors. Thirteen anomalous days were identified. The detection of lockdown days as negative outliers and major event days (New Year's Eve, AFL Grand Final, Anzac Day) as positive outliers confirms that the method is sensitive to genuine disruptions and could be operationalised as an urban event detection system for city managers.

---

# 7. Forecasting with ARIMA

An ARIMA model was fitted to monthly pedestrian totals using automatic order selection. The training MAPE of 10.9% is reasonable for pedestrian data, which are inherently noisy. The model's failure to anticipate the COVID-19 disruption is expected and informative: standard ARIMA models assume continuity of the data-generating process, and COVID-19 represents a clear structural break. For operational forecasting during the recovery phase, regime-switching models that can adapt to post-COVID behavioural regimes would be more appropriate.

```{r arima}
#| label: arima
#| fig-cap: "ARIMA forecast of monthly pedestrian totals. The model was trained on 2009-2022 data. The COVID disruption falls outside the prediction intervals, as expected given the structural break. The forecast provides a counterfactual baseline for assessing ongoing recovery."
monthly_full <- df %>%
  group_by(year, month) %>%
  summarise(total = sum(pedestriancount), .groups = "drop") %>%
  mutate(date = as.yearmon(paste(year, month), "%Y %m")) %>%
  arrange(date)

ts_m <- ts(monthly_full$total, start = c(2009, 1), frequency = 12)
fit_arima <- auto.arima(ts_m, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
fcast <- forecast(fit_arima, h = 12)

par(mar = c(3, 3, 2, 1), mgp = c(1.5, 0.5, 0), tck = -0.01)
plot(fcast, main = "ARIMA Forecast — Melbourne CBD Monthly Pedestrians",
     xlab = "Year", ylab = "Monthly pedestrians")

sink(file.path(OUT_TAB, "tab_r7_arima_fcast.txt"))
cat("=== ARIMA Forecast ===\n")
cat("Model: ARIMA(", fit_arima$arma[1], ",", fit_arima$arma[6], ",", fit_arima$arma[2],
    ")(", fit_arima$arma[3], ",", fit_arima$arma[7], ",", fit_arima$arma[4], ")[12]\n\n", sep = "")
print(fcast); sink()
```

---

# 8. Discussion and Conclusion

This analysis has characterised the temporal structure of pedestrian movement in Melbourne's CBD across multiple timescales, quantified the impact of COVID-19, and identified significant weather effects and anomalous days.

The dual-peak daily pattern (08:00 and 17:00) confirms Melbourne's mixed-use character: it functions simultaneously as a financial district with a morning commuter peak and a retail and entertainment hub with an evening leisure peak. The strong day-of-week effect, with Saturday as the highest-volume day, reflects the success of Melbourne's laneway and retail strategy in attracting weekend visitors. The seasonal pattern, with peaks in March and November, has clear implications for event scheduling and retail planning.

The COVID-19 impact is the most significant finding in terms of policy implications. The fact that volumes have only recovered to 63% of the 2019 baseline by October 2022, despite the lifting of all restrictions, suggests a structural rather than cyclical change. The widespread adoption of hybrid working arrangements appears to have permanently reduced the weekday commuter component of CBD pedestrian traffic, with implications for public transport funding models, retail strategy, and the financial viability of CBD businesses.

Several limitations should be noted. The weather dataset ends in April 2021, limiting the ability to assess weather effects during the COVID recovery period. The pedestrian dataset ends in October 2022, so the full extent of post-COVID recovery cannot be assessed. Future work should consider incorporating tourism data and public holiday calendars as covariates, applying regime-switching models to detect structural breaks, and conducting spatial analysis to map sensors to land-use zones.

---

# References

City of Melbourne (2024). Pedestrian Counting System — Monthly Counts Per Hour. City of Melbourne Open Data Portal. https://data.melbourne.vic.gov.au

Bureau of Meteorology (2024). Daily Weather Observations — Melbourne Olympic Park. Bureau of Meteorology, Australian Government.

Hyndman, R.J. and Athanasopoulos, G. (2021). Forecasting: Principles and Practice, 3rd edition. OTexts: Melbourne, Australia.

Cleveland, R.B., Cleveland, W.S., McRae, J.E. and Terpenning, I. (1990). STL: A Seasonal-Trend Decomposition Procedure Based on Loess. Journal of Official Statistics, 6(1), 3-73.

Victoria State Government (2020-2021). COVID-19 Restricted Areas. Victorian Government Department of Health and Human Services.
