---
title: "analysis_draft"
format: html
editor: visual
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r}
# Loading all the libraries
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(fs)
library(httr2)
library(arrow)
library(here)
library(ggrepel)
library(plotly)
library(leaflet)
```

```{r}
dir_create("Data_pedestrians/raw2")

raw_path <- "Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrian_url <- paste0(
  "https://data.melbourne.vic.gov.au/",
  "api/explore/v2.1/catalog/datasets/",
  "pedestrian-counting-system-monthly-counts-per-hour/",
  "exports/csv?delimiter=%2C&with_bom=false"
)

request(pedestrian_url) |>
  req_perform(path = raw_path)
```

```{r}
#Historical data
historical <- read_csv(
  here(
    "Data_pedestrians",
    "raw",
    "Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv"
  ),
  show_col_types = FALSE
) |>
  clean_names()


#Current API data
current <- read_csv(
  here(
    "Data_pedestrians",
    "raw2",
    "pedestrian_counts_raw.csv"
  ),
  show_col_types = FALSE
) |>
  clean_names()

```

```{r}
#Historical data
historical_clean <- historical |>
  mutate(
    sensing_date = mdy(
      str_c(month, mdate, year, sep = " ")
    )
  ) |>
  group_by(
    sensor_id,
    sensing_date
  ) |>
  arrange(
    id,
    .by_group = TRUE
  ) |>
  mutate(
    hour_repaired =
      n() == 24 & n_distinct(time) == 1,

    hourday = if_else(
      hour_repaired,
      row_number() - 1,
      time
    )
  ) |>
  ungroup() |>
  transmute(
    location_id = sensor_id,
    sensing_date,
    hourday,
    pedestrian_count = hourly_counts,
    hour_repaired,
    source = "Historical"
  )

#Current API data
current_clean <- current |>
  transmute(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count = pedestriancount,
    source = "Current API"
  )

#Combine
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) |>
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

```{r}
pedestrian_full <- pedestrian_full |>
  mutate(
    year = year(sensing_date),
    month = month(sensing_date),
    month_name = month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    ),
    day_of_week = wday(
      sensing_date,
      week_start = 1
    ),
    is_weekend = day_of_week >= 6,
    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  )
```

```{r}
#Check for missing values
pedestrian_full |>
  summarise(
    observations = n(),
    locations = n_distinct(location_id),
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    missing_dates = sum(is.na(sensing_date)),
    missing_hours = sum(is.na(hourday)),
    missing_counts = sum(is.na(pedestrian_count))
  )

pedestrian_full |>
  summarise(
    negative_counts = sum(pedestrian_count < 0, na.rm = TRUE),
    invalid_hours = sum(
      hourday < 0 | hourday > 23,
      na.rm = TRUE
    )
  )

#Check for duplicates
duplicates <- pedestrian_full |>
  count(
    location_id,
    sensing_date,
    hourday,
    name = "n"
  ) |>
  filter(n > 1)

duplicates
```

-   A small number of historical sensor-day groups contained 24 distinct hourly count observations but duplicated timestamps; where the record ordering clearly preserved the hourly sequence, hour labels were reconstructed from record order, with repaired observations flagged.

```{r}
#Download  the sensors location dataset

sensor_locations <- read_csv(
  paste0(
    "https://data.melbourne.vic.gov.au/",
    "api/explore/v2.1/catalog/datasets/",
    "pedestrian-counting-system-sensor-locations/",
    "exports/csv?delimiter=%2C"
  ),
  show_col_types = FALSE
) |>
  clean_names()

sensor_locations |>
  glimpse()
```

-   Has the pedestrian monitoring network changed enough over time to affect comparisons? how many sensor locations were actually contributing usable pedestrian data in each year

```{r}
year_window <- pedestrian_full |>
  group_by(year) |>
  summarise(
    available_days = n_distinct(sensing_date),
    .groups = "drop"
  )

sensor_year_coverage <- pedestrian_full |>
  group_by(
    year,
    location_id
  ) |>
  summarise(
    days_observed = n_distinct(sensing_date),
    observations = n(),
    .groups = "drop"
  ) |>
  left_join(
    year_window,
    by = "year"
  ) |>
  mutate(
    coverage = days_observed / available_days
  )
```

```{r}
year_coverage <- sensor_year_coverage |>
  group_by(year) |>
  summarise(
    active_locations = n_distinct(location_id),
    well_covered_locations = sum(coverage >= 0.80),
    .groups = "drop"
  ) |>
  complete(
    year = full_seq(year, 1)
  ) |>
  pivot_longer(
    cols = c(
      active_locations,
      well_covered_locations
    ),
    names_to = "coverage_type",
    values_to = "locations"
  ) |>
  mutate(
    coverage_type = recode(
      coverage_type,
      active_locations = "Any observations",
      well_covered_locations = "At least 80% coverage"
    )
  )

year_coverage |>
  ggplot(
    aes(
      x = year,
      y = locations,
      linetype = coverage_type
    )
  ) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  labs(
    title = "Melbourne's pedestrian monitoring network change over time",
    subtitle = "2023 is left as a genuine data gap",
    x = "Year",
    y = "Number of sensor locations",
    linetype = NULL
  ) +
  theme_minimal()
```

-   When were the currently documented sensors installed? How many new sensors were installed in each year?

```{r}
sensor_installations <- sensor_locations |>
  filter(!is.na(installation_date)) |>
  mutate(
    installation_year = year(installation_date)
  ) |>
  count(
    installation_year,
    name = "new_sensors"
  )

sensor_installations |>
  ggplot(
    aes(
      x = installation_year,
      y = new_sensors
    )
  ) +
  geom_col() +
  labs(
    title = "Expansion of Melbourne's pedestrian sensor network",
    x = "Installation year",
    y = "Sensors installed"
  ) +
  theme_minimal()
```

-   Is there a monthly or seasonal rhythm?

```{r}
# Monthly rhythm using the full available dataset: 2009–2026
monthly_profile <- pedestrian_full |>
  filter(
    year >= 2009,
    year <= 2026
  ) |>
  group_by(
    location_id,
    sensing_date,
    month,
    month_name
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    location_id,
    month,
    month_name
  ) |>
  summarise(
    sensor_month_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    month,
    month_name
  ) |>
  summarise(
    mean_daily_count = mean(
      sensor_month_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(month)
```

```{r}
monthly_profile |>
  ggplot(
    aes(
      x = month,
      y = mean_daily_count
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = monthly_profile$month,
    labels = monthly_profile$month_name
  ) +
  labs(
    title = "Melbourne's monthly pedestrian rhythm",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

-   seasons

```{r}
seasonal_profile <- pedestrian_full |>
  
  # Step 1: Daily total for each sensor
  group_by(
    location_id,
    year,
    sensing_date,
    season
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 2: Typical activity for each sensor in each season/year
  group_by(
    location_id,
    year,
    season
  ) |>
  summarise(
    sensor_season_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 3: Typical city activity within each year and season
  group_by(
    year,
    season
  ) |>
  summarise(
    yearly_season_mean = mean(
      sensor_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 4: Average seasonal pattern across all available years
  group_by(season) |>
  summarise(
    mean_daily_count = mean(
      yearly_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  mutate(
    season = fct_relevel(
      season,
      "Summer",
      "Autumn",
      "Winter",
      "Spring"
    )
  )
```

```{r}
seasonal_profile |>
  ggplot(
    aes(
      x = season,
      y = mean_daily_count
    )
  ) +
  geom_col() +
  labs(
    title = "Seasonal pattern in Melbourne pedestrian activity",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal()
```

-   Covid disruption

```{r}

covid_daily <- pedestrian_full |>
  filter(
    source == "Historical",
    year >= 2015,
    year <= 2022
  ) |>
  group_by(
    location_id,
    sensing_date,
    year,
    month
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    month_date = floor_date(
      sensing_date,
      "month"
    )
  )
```

```{r}
sensor_monthly <- covid_daily |>
  group_by(
    location_id,
    month_date,
    month
  ) |>
  summarise(
    mean_daily_count = mean(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n_distinct(sensing_date),
    .groups = "drop"
  ) |>
  filter(
    days_observed >= 20
  )
```

```{r}
pre_covid_baseline <- sensor_monthly |>
  filter(
    year(month_date) >= 2015,
    year(month_date) <= 2019
  ) |>
  group_by(
    location_id,
    month
  ) |>
  summarise(
    baseline_count = mean(
      mean_daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

```{r}
covid_index <- sensor_monthly |>
  left_join(
    pre_covid_baseline,
    by = c(
      "location_id",
      "month"
    )
  ) |>
  filter(
    !is.na(baseline_count),
    baseline_count > 0
  ) |>
  mutate(
    activity_index =
      mean_daily_count / baseline_count * 100
  )
```

```{r}
melbourne_covid_index <- covid_index |>
  group_by(month_date) |>
  summarise(
    activity_index = median(
      activity_index,
      na.rm = TRUE
    ),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

```{r}
melbourne_covid_index |>
  ggplot(
    aes(
      x = month_date,
      y = activity_index
    )
  ) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  geom_line(
    linewidth = 0.9
  ) +
  labs(
    title = "COVID-19 disrupted Melbourne's pedestrian pulse",
    subtitle = "Monthly activity relative to the same-month 2015–2019 baseline",
    x = NULL,
    y = "Pedestrian activity index\n(100 = pre-COVID baseline)"
  ) +
  theme_minimal()
```


- Where is the Melbourne's pulse the strongest currently?
```{r}
location_metadata <- sensor_locations |>
  mutate(
    relocation_flag = str_detect(
      str_to_lower(
        coalesce(note, "")
      ),
      "relocat|moved to|move to"
    )
  ) |>
  filter(
    !is.na(latitude),
    !is.na(longitude)
  ) |>
  select(
    location_id,
    sensor_description,
    sensor_name,
    latitude,
    longitude,
    installation_date,
    note,
    relocation_flag
  )
```

```{r}
sensor_daily_spatial <- pedestrian_full |>
  group_by(
    location_id,
    sensing_date,
    year,
    month
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    hours_observed = n_distinct(hourday),
    .groups = "drop"
  ) |>
  filter(
    hours_observed >= 20
  )
```

```{r}
recent_strength <- sensor_daily_spatial |>
  filter(
    year == 2025
  ) |>
  group_by(location_id) |>
  summarise(
    typical_daily_count = median(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n(),
    .groups = "drop"
  ) |>
  filter(
    days_observed >= 292
  ) |>
  inner_join(
    location_metadata,
    by = "location_id"
  )
```

```{r}

spatial_plot <- recent_strength |>
  ggplot(
    aes(
      x = longitude,
      y = latitude,
      size = typical_daily_count,
      text = paste0(
        "<b>", sensor_description, "</b>",
        "<br>Location ID: ", location_id,
        "<br>Typical daily count: ",
        scales::comma(round(typical_daily_count))
      )
    )
  ) +
  geom_point(alpha = 0.65) +
  coord_quickmap() +
  labs(
    title = "Where is Melbourne's pedestrian pulse strongest?",
    subtitle = "Typical daily pedestrian activity at well-covered sensor locations, 2025",
    x = NULL,
    y = NULL,
    size = "Typical daily count"
  ) +
  theme_minimal()

ggplotly(
  spatial_plot,
  tooltip = "text"
)
```
```{r}

top_10_locations |>
  mutate(
    sensor_description = fct_reorder(
      sensor_description,
      typical_daily_count
    )
  ) |>
  ggplot(
    aes(
      x = sensor_description,
      y = typical_daily_count
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Melbourne's busiest monitored pedestrian locations",
    subtitle = "Typical daily pedestrian activity, 2025",
    x = NULL,
    y = "Typical daily pedestrian count"
  ) +
  theme_minimal()
```

- How does recent pedestrian activity vary across Melbourne locations relative to their pre-COVID patterns?

```{r}
sensor_daily_change <- pedestrian_full |>
  group_by(
    location_id,
    sensing_date,
    year,
    month
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    hours_observed = n_distinct(hourday),
    .groups = "drop"
  ) |>
  filter(
    hours_observed >= 20
  )
```

```{r}
sensor_monthly_change <- sensor_daily_change |>
  mutate(
    month_date = floor_date(
      sensing_date,
      "month"
    )
  ) |>
  group_by(
    location_id,
    year,
    month,
    month_date
  ) |>
  summarise(
    mean_daily_count = mean(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n(),
    .groups = "drop"
  ) |>
  filter(
    days_observed >= 20
  )
```

```{r}
pre_covid_location_baseline <- sensor_monthly_change |>
  filter(
    year >= 2015,
    year <= 2019
  ) |>
  group_by(
    location_id,
    month
  ) |>
  summarise(
    baseline_count = median(
      mean_daily_count,
      na.rm = TRUE
    ),
    baseline_years = n_distinct(year),
    .groups = "drop"
  ) |>
  filter(
    baseline_years >= 3
  )
```

```{r}
recent_location_change <- sensor_monthly_change |>
  filter(
    year >= 2024,
    year <= 2026
  ) |>
  inner_join(
    pre_covid_location_baseline,
    by = c(
      "location_id",
      "month"
    )
  ) |>
  mutate(
    activity_index =
      mean_daily_count / baseline_count * 100
  )
```


```{r}
location_recovery_robust <- recent_location_change |>
  group_by(location_id) |>
  summarise(
    recovery_index = median(
      activity_index,
      na.rm = TRUE
    ),
    recent_months = n_distinct(month_date),
    .groups = "drop"
  ) |>
  filter(recent_months >= 20) |>
  mutate(
    change_percent = recovery_index - 100
  )

location_recovery_robust |>
  summarise(
    locations = n(),
    median_change = median(
      change_percent,
      na.rm = TRUE
    ),
    q25 = quantile(
      change_percent,
      0.25,
      na.rm = TRUE
    ),
    q75 = quantile(
      change_percent,
      0.75,
      na.rm = TRUE
    ),
    minimum_change = min(
      change_percent,
      na.rm = TRUE
    ),
    maximum_change = max(
      change_percent,
      na.rm = TRUE
    )
  )
```

```{r}
location_recovery_final <- location_recovery_robust |>
  inner_join(
    location_metadata,
    by = "location_id"
  ) |>
  filter(
    !is.na(latitude),
    !is.na(longitude),
    !relocation_flag
  )

location_recovery_final |>
  summarise(
    locations = n_distinct(location_id)
  )
```


```{r}

recovery_plot <- location_recovery_final |>
  ggplot(
    aes(
      x = longitude,
      y = latitude,
      colour = change_percent,
      text = paste0(
        "<b>", sensor_description, "</b>",
        "<br>Location ID: ", location_id,
        "<br>Recent activity index: ",
        round(recovery_index, 1),
        "<br>Change from pre-COVID: ",
        round(change_percent, 1), "%",
        "<br>Recent months available: ",
        recent_months
      )
    )
  ) +
  geom_point(
    size = 4,
    alpha = 0.8
  ) +
  coord_quickmap() +
  scale_colour_gradient2(
    midpoint = 0
  ) +
  labs(
    title = "Recent pedestrian activity varies across Melbourne locations",
    subtitle = "Activity relative to location-specific, same-month 2015–2019 baselines",
    x = NULL,
    y = NULL,
    colour = "Change (%)"
  ) +
  theme_minimal()



ggplotly(
  recovery_plot,
  tooltip = "text"
)
```

- How did each location change after the covid?

- Does the extreme −72.6% change at Flinders Underpass represent a reliable behavioural shift, or is long-term comparability at this sensor uncertain?

```{r}
flinders_metadata <- sensor_locations |>
  filter(location_id == 6) |>
  select(
    location_id,
    sensor_description,
    sensor_name,
    installation_date,
    status,
    note
  )

flinders_metadata
```

```{r}
flinders_monthly <- pedestrian_full |>
  filter(location_id == 6) |>
  group_by(
    source,
    sensing_date
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    hours_observed = n_distinct(hourday),
    .groups = "drop"
  ) |>
  filter(
    hours_observed >= 20
  ) |>
  mutate(
    month_date = floor_date(
      sensing_date,
      "month"
    )
  ) |>
  group_by(
    source,
    month_date
  ) |>
  summarise(
    mean_daily_count = mean(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n(),
    .groups = "drop"
  )
```

```{r}
flinders_monthly |>
  ggplot(
    aes(
      x = month_date,
      y = mean_daily_count,
      group = source,
      linetype = source
    )
  ) +
  geom_line() +
  geom_vline(
    xintercept = ymd("2021-09-08"),
    linetype = "dashed"
  ) +
  labs(
    title = "Pedestrian activity at Flinders Underpass – Myki Barriers",
    subtitle = "Dashed vertical line marks the recorded sensor upgrade on 8 September 2021",
    x = NULL,
    y = "Average daily pedestrian count",
    linetype = "Data source"
  ) +
  theme_minimal()
```

```{r}
# Inspecting the period around the upgrade
flinders_monthly |>
  filter(
    month_date >= ymd("2021-01-01"),
    month_date <= ymd("2022-10-01")
  ) |>
  ggplot(
    aes(
      x = month_date,
      y = mean_daily_count
    )
  ) +
  geom_line() +
  geom_point() +
  geom_vline(
    xintercept = ymd("2021-09-08"),
    linetype = "dashed"
  ) +
  labs(
    title = "Flinders Underpass activity around the 2021 sensor upgrade",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal()
```
- There is no clear evidence that the September 2021 upgrade itself caused the long-term decline.

```{r}
flinders_source_summary <- pedestrian_full |>
  filter(location_id == 6) |>
  group_by(source) |>
  summarise(
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    observations = n(),
    .groups = "drop"
  )

flinders_source_summary
```

```{r}
# Check whether the result is driven by this one sensor
spatial_with_flinders <- location_recovery_final |>
  summarise(
    locations = n(),
    median_change = median(
      change_percent,
      na.rm = TRUE
    ),
    q25 = quantile(
      change_percent,
      0.25,
      na.rm = TRUE
    ),
    q75 = quantile(
      change_percent,
      0.75,
      na.rm = TRUE
    )
  ) |>
  mutate(
    analysis = "Including Flinders Underpass"
  )

location_recovery_no_flinders <- location_recovery_final |>
  filter(location_id != 6)

spatial_without_flinders <- location_recovery_no_flinders |>
  summarise(
    locations = n(),
    median_change = median(
      change_percent,
      na.rm = TRUE
    ),
    q25 = quantile(
      change_percent,
      0.25,
      na.rm = TRUE
    ),
    q75 = quantile(
      change_percent,
      0.75,
      na.rm = TRUE
    )
  ) |>
  mutate(
    analysis = "Excluding Flinders Underpass"
  )

flinders_sensitivity <- bind_rows(
  spatial_with_flinders,
  spatial_without_flinders
) |>
  select(
    analysis,
    locations,
    median_change,
    q25,
    q75
  )

flinders_sensitivity
```








