---
title: "analysis_draft"
format: html
editor: visual
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r}
# Loading all the libraries
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(fs)
library(httr2)
library(arrow)
library(here)
```

```{r}
dir_create("Data_pedestrians/raw2")

raw_path <- "Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrian_url <- paste0(
  "https://data.melbourne.vic.gov.au/",
  "api/explore/v2.1/catalog/datasets/",
  "pedestrian-counting-system-monthly-counts-per-hour/",
  "exports/csv?delimiter=%2C&with_bom=false"
)

request(pedestrian_url) |>
  req_perform(path = raw_path)
```

```{r}
#Historical data
historical <- read_csv(
  here(
    "Data_pedestrians",
    "raw",
    "Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv"
  ),
  show_col_types = FALSE
) |>
  clean_names()


#Current API data
current <- read_csv(
  here(
    "Data_pedestrians",
    "raw2",
    "pedestrian_counts_raw.csv"
  ),
  show_col_types = FALSE
) |>
  clean_names()

```

```{r}
#Historical data
historical_clean <- historical |>
  mutate(
    sensing_date = mdy(
      str_c(month, mdate, year, sep = " ")
    )
  ) |>
  group_by(
    sensor_id,
    sensing_date
  ) |>
  arrange(
    id,
    .by_group = TRUE
  ) |>
  mutate(
    hour_repaired =
      n() == 24 & n_distinct(time) == 1,

    hourday = if_else(
      hour_repaired,
      row_number() - 1,
      time
    )
  ) |>
  ungroup() |>
  transmute(
    location_id = sensor_id,
    sensing_date,
    hourday,
    pedestrian_count = hourly_counts,
    hour_repaired,
    source = "Historical"
  )

#Current API data
current_clean <- current |>
  transmute(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count = pedestriancount,
    source = "Current API"
  )

#Combine
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) |>
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

```{r}
pedestrian_full <- pedestrian_full |>
  mutate(
    year = year(sensing_date),
    month = month(sensing_date),
    month_name = month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    ),
    day_of_week = wday(
      sensing_date,
      week_start = 1
    ),
    is_weekend = day_of_week >= 6,
    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  )
```

```{r}
#Check for missing values
pedestrian_full |>
  summarise(
    observations = n(),
    locations = n_distinct(location_id),
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    missing_dates = sum(is.na(sensing_date)),
    missing_hours = sum(is.na(hourday)),
    missing_counts = sum(is.na(pedestrian_count))
  )

pedestrian_full |>
  summarise(
    negative_counts = sum(pedestrian_count < 0, na.rm = TRUE),
    invalid_hours = sum(
      hourday < 0 | hourday > 23,
      na.rm = TRUE
    )
  )

#Check for duplicates
duplicates <- pedestrian_full |>
  count(
    location_id,
    sensing_date,
    hourday,
    name = "n"
  ) |>
  filter(n > 1)

duplicates
```

- A small number of historical sensor-day groups contained 24 distinct hourly count observations but duplicated timestamps; where the record ordering clearly preserved the hourly sequence, hour labels were reconstructed from record order, with repaired observations flagged.

```{r}
#Download  the sensors location dataset

sensor_locations <- read_csv(
  paste0(
    "https://data.melbourne.vic.gov.au/",
    "api/explore/v2.1/catalog/datasets/",
    "pedestrian-counting-system-sensor-locations/",
    "exports/csv?delimiter=%2C"
  ),
  show_col_types = FALSE
) |>
  clean_names()

sensor_locations |>
  glimpse()
```

- Has the pedestrian monitoring network changed enough over time to affect comparisons? how many sensor locations were actually contributing usable pedestrian data in each year
```{r}
year_window <- pedestrian_full |>
  group_by(year) |>
  summarise(
    available_days = n_distinct(sensing_date),
    .groups = "drop"
  )

sensor_year_coverage <- pedestrian_full |>
  group_by(
    year,
    location_id
  ) |>
  summarise(
    days_observed = n_distinct(sensing_date),
    observations = n(),
    .groups = "drop"
  ) |>
  left_join(
    year_window,
    by = "year"
  ) |>
  mutate(
    coverage = days_observed / available_days
  )
```

```{r}
year_coverage <- sensor_year_coverage |>
  group_by(year) |>
  summarise(
    active_locations = n_distinct(location_id),
    well_covered_locations = sum(coverage >= 0.80),
    .groups = "drop"
  ) |>
  complete(
    year = full_seq(year, 1)
  ) |>
  pivot_longer(
    cols = c(
      active_locations,
      well_covered_locations
    ),
    names_to = "coverage_type",
    values_to = "locations"
  ) |>
  mutate(
    coverage_type = recode(
      coverage_type,
      active_locations = "Any observations",
      well_covered_locations = "At least 80% coverage"
    )
  )

year_coverage |>
  ggplot(
    aes(
      x = year,
      y = locations,
      linetype = coverage_type
    )
  ) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  labs(
    title = "Melbourne's pedestrian monitoring network change over time",
    subtitle = "2023 is left as a genuine data gap",
    x = "Year",
    y = "Number of sensor locations",
    linetype = NULL
  ) +
  theme_minimal()
```

- When were the currently documented sensors installed? How many new sensors were installed in each year?

```{r}
sensor_installations <- sensor_locations |>
  filter(!is.na(installation_date)) |>
  mutate(
    installation_year = year(installation_date)
  ) |>
  count(
    installation_year,
    name = "new_sensors"
  )

sensor_installations |>
  ggplot(
    aes(
      x = installation_year,
      y = new_sensors
    )
  ) +
  geom_col() +
  labs(
    title = "Expansion of Melbourne's pedestrian sensor network",
    x = "Installation year",
    y = "Sensors installed"
  ) +
  theme_minimal()
```

- Is there a monthly or seasonal rhythm?

```{r}
# Monthly rhythm using the full available dataset: 2009–2026
monthly_profile <- pedestrian_full |>
  filter(
    year >= 2009,
    year <= 2026
  ) |>
  group_by(
    location_id,
    sensing_date,
    month,
    month_name
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    location_id,
    month,
    month_name
  ) |>
  summarise(
    sensor_month_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    month,
    month_name
  ) |>
  summarise(
    mean_daily_count = mean(
      sensor_month_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(month)
```

```{r}
monthly_profile |>
  ggplot(
    aes(
      x = month,
      y = mean_daily_count
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = monthly_profile$month,
    labels = monthly_profile$month_name
  ) +
  labs(
    title = "Melbourne's monthly pedestrian rhythm",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

- seasons

```{r}
seasonal_profile <- pedestrian_full |>
  
  # Step 1: Daily total for each sensor
  group_by(
    location_id,
    year,
    sensing_date,
    season
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 2: Typical activity for each sensor in each season/year
  group_by(
    location_id,
    year,
    season
  ) |>
  summarise(
    sensor_season_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 3: Typical city activity within each year and season
  group_by(
    year,
    season
  ) |>
  summarise(
    yearly_season_mean = mean(
      sensor_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 4: Average seasonal pattern across all available years
  group_by(season) |>
  summarise(
    mean_daily_count = mean(
      yearly_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  mutate(
    season = fct_relevel(
      season,
      "Summer",
      "Autumn",
      "Winter",
      "Spring"
    )
  )
```

```{r}
seasonal_profile |>
  ggplot(
    aes(
      x = season,
      y = mean_daily_count
    )
  ) +
  geom_col() +
  labs(
    title = "Seasonal pattern in Melbourne pedestrian activity",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal()
```

- Covid disruption

```{r}

covid_daily <- pedestrian_full |>
  filter(
    source == "Historical",
    year >= 2015,
    year <= 2022
  ) |>
  group_by(
    location_id,
    sensing_date,
    year,
    month
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    month_date = floor_date(
      sensing_date,
      "month"
    )
  )
```

```{r}
sensor_monthly <- covid_daily |>
  group_by(
    location_id,
    month_date,
    month
  ) |>
  summarise(
    mean_daily_count = mean(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n_distinct(sensing_date),
    .groups = "drop"
  ) |>
  filter(
    days_observed >= 20
  )
```

```{r}
pre_covid_baseline <- sensor_monthly |>
  filter(
    year(month_date) >= 2015,
    year(month_date) <= 2019
  ) |>
  group_by(
    location_id,
    month
  ) |>
  summarise(
    baseline_count = mean(
      mean_daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

```{r}
covid_index <- sensor_monthly |>
  left_join(
    pre_covid_baseline,
    by = c(
      "location_id",
      "month"
    )
  ) |>
  filter(
    !is.na(baseline_count),
    baseline_count > 0
  ) |>
  mutate(
    activity_index =
      mean_daily_count / baseline_count * 100
  )
```

```{r}
melbourne_covid_index <- covid_index |>
  group_by(month_date) |>
  summarise(
    activity_index = median(
      activity_index,
      na.rm = TRUE
    ),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

```{r}
melbourne_covid_index |>
  ggplot(
    aes(
      x = month_date,
      y = activity_index
    )
  ) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  geom_line(
    linewidth = 0.9
  ) +
  labs(
    title = "COVID-19 disrupted Melbourne's pedestrian pulse",
    subtitle = "Monthly activity relative to the same-month 2015–2019 baseline",
    x = NULL,
    y = "Pedestrian activity index\n(100 = pre-COVID baseline)"
  ) +
  theme_minimal()
```
