---
title: "analysis_draft"
format: html
editor: visual
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r}
# Loading all the libraries
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(fs)
library(httr2)
library(arrow)
```

```{r}
dir_create("Data_pedestrians/raw2")

raw_path <- "Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrian_url <- paste0(
  "https://data.melbourne.vic.gov.au/",
  "api/explore/v2.1/catalog/datasets/",
  "pedestrian-counting-system-monthly-counts-per-hour/",
  "exports/csv?delimiter=%2C&with_bom=false"
)

request(pedestrian_url) |>
  req_perform(path = raw_path)
```

```{r}
#Historical data
historical <- read_csv(
  "../Data_pedestrians/raw/Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv",
  show_col_types = FALSE
) |>
  clean_names()

#Current API data
current <- read_csv(
  "../Data_pedestrians/raw2/pedestrian_counts_raw.csv",
  show_col_types = FALSE
) |>
  clean_names()

```

```{r}
#Historical data
historical_clean <- historical |>
  mutate(
    sensing_date = mdy(
      str_c(month, mdate, year, sep = " ")
    )
  ) |>
  group_by(
    sensor_id,
    sensing_date
  ) |>
  arrange(
    id,
    .by_group = TRUE
  ) |>
  mutate(
    hour_repaired =
      n() == 24 & n_distinct(time) == 1,

    hourday = if_else(
      hour_repaired,
      row_number() - 1,
      time
    )
  ) |>
  ungroup() |>
  transmute(
    location_id = sensor_id,
    sensing_date,
    hourday,
    pedestrian_count = hourly_counts,
    hour_repaired,
    source = "Historical"
  )

#Current API data
current_clean <- current |>
  transmute(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count = pedestriancount,
    source = "Current API"
  )

#Combine
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) |>
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

```{r}
pedestrian_full <- pedestrian_full |>
  mutate(
    year = year(sensing_date),
    month = month(sensing_date),
    month_name = month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    ),
    day_of_week = wday(
      sensing_date,
      week_start = 1
    ),
    is_weekend = day_of_week >= 6,
    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  )
```

```{r}
#Check for missing values
pedestrian_full |>
  summarise(
    observations = n(),
    locations = n_distinct(location_id),
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    missing_dates = sum(is.na(sensing_date)),
    missing_hours = sum(is.na(hourday)),
    missing_counts = sum(is.na(pedestrian_count))
  )

pedestrian_full |>
  summarise(
    negative_counts = sum(pedestrian_count < 0, na.rm = TRUE),
    invalid_hours = sum(
      hourday < 0 | hourday > 23,
      na.rm = TRUE
    )
  )

#Check for duplicates
duplicates <- pedestrian_full |>
  count(
    location_id,
    sensing_date,
    hourday,
    name = "n"
  ) |>
  filter(n > 1)

duplicates
```

-   A small number of historical sensor-day groups contained 24 distinct hourly count observations but duplicated timestamps; where the record ordering clearly preserved the hourly sequence, hour labels were reconstructed from record order, with repaired observations flagged.

```{r}
#Download  the sensors location dataset

sensor_locations <- read_csv(
  paste0(
    "https://data.melbourne.vic.gov.au/",
    "api/explore/v2.1/catalog/datasets/",
    "pedestrian-counting-system-sensor-locations/",
    "exports/csv?delimiter=%2C"
  ),
  show_col_types = FALSE
) |>
  clean_names()

sensor_locations |>
  glimpse()
```

-   Has the pedestrian monitoring network changed enough over time to affect comparisons? how many sensor locations were actually contributing usable pedestrian data in each year

```{r}
year_window <- pedestrian_full |>
  group_by(year) |>
  summarise(
    available_days = n_distinct(sensing_date),
    .groups = "drop"
  )

sensor_year_coverage <- pedestrian_full |>
  group_by(
    year,
    location_id
  ) |>
  summarise(
    days_observed = n_distinct(sensing_date),
    observations = n(),
    .groups = "drop"
  ) |>
  left_join(
    year_window,
    by = "year"
  ) |>
  mutate(
    coverage = days_observed / available_days
  )
```

```{r}
year_coverage <- sensor_year_coverage |>
  group_by(year) |>
  summarise(
    active_locations = n_distinct(location_id),
    well_covered_locations = sum(coverage >= 0.80),
    .groups = "drop"
  ) |>
  complete(
    year = full_seq(year, 1)
  ) |>
  pivot_longer(
    cols = c(
      active_locations,
      well_covered_locations
    ),
    names_to = "coverage_type",
    values_to = "locations"
  ) |>
  mutate(
    coverage_type = recode(
      coverage_type,
      active_locations = "Any observations",
      well_covered_locations = "At least 80% coverage"
    )
  )

year_coverage |>
  ggplot(
    aes(
      x = year,
      y = locations,
      linetype = coverage_type
    )
  ) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  labs(
    title = "Melbourne's pedestrian monitoring network change over time",
    subtitle = "2023 is left as a genuine data gap",
    x = "Year",
    y = "Number of sensor locations",
    linetype = NULL
  ) +
  theme_minimal()
```

-   When were the currently documented sensors installed? How many new sensors were installed in each year?

```{r}
sensor_installations <- sensor_locations |>
  filter(!is.na(installation_date)) |>
  mutate(
    installation_year = year(installation_date)
  ) |>
  count(
    installation_year,
    name = "new_sensors"
  )

sensor_installations |>
  ggplot(
    aes(
      x = installation_year,
      y = new_sensors
    )
  ) +
  geom_col() +
  labs(
    title = "Expansion of Melbourne's pedestrian sensor network",
    x = "Installation year",
    y = "Sensors installed"
  ) +
  theme_minimal()
```

-   Is there a monthly or seasonal rhythm?

```{r}
# Monthly rhythm using the full available dataset: 2009–2026
monthly_profile <- pedestrian_full |>
  filter(
    year >= 2009,
    year <= 2026
  ) |>
  group_by(
    location_id,
    sensing_date,
    month,
    month_name
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    location_id,
    month,
    month_name
  ) |>
  summarise(
    sensor_month_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  group_by(
    month,
    month_name
  ) |>
  summarise(
    mean_daily_count = mean(
      sensor_month_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(month)
```

```{r}
monthly_profile |>
  ggplot(
    aes(
      x = month,
      y = mean_daily_count
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = monthly_profile$month,
    labels = monthly_profile$month_name
  ) +
  labs(
    title = "Melbourne's monthly pedestrian rhythm",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

-   seasons

```{r}
seasonal_profile <- pedestrian_full |>
  
  # Step 1: Daily total for each sensor
  group_by(
    location_id,
    year,
    sensing_date,
    season
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 2: Typical activity for each sensor in each season/year
  group_by(
    location_id,
    year,
    season
  ) |>
  summarise(
    sensor_season_mean = mean(
      daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 3: Typical city activity within each year and season
  group_by(
    year,
    season
  ) |>
  summarise(
    yearly_season_mean = mean(
      sensor_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  # Step 4: Average seasonal pattern across all available years
  group_by(season) |>
  summarise(
    mean_daily_count = mean(
      yearly_season_mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  
  mutate(
    season = fct_relevel(
      season,
      "Summer",
      "Autumn",
      "Winter",
      "Spring"
    )
  )
```

```{r}
seasonal_profile |>
  ggplot(
    aes(
      x = season,
      y = mean_daily_count
    )
  ) +
  geom_col() +
  labs(
    title = "Seasonal pattern in Melbourne pedestrian activity",
    subtitle = "Average daily pedestrian activity across available data, 2009–2026",
    x = NULL,
    y = "Average daily pedestrian count"
  ) +
  theme_minimal()
```

-   Covid disruption

```{r}

covid_daily <- pedestrian_full |>
  filter(
    source == "Historical",
    year >= 2015,
    year <= 2022
  ) |>
  group_by(
    location_id,
    sensing_date,
    year,
    month
  ) |>
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    month_date = floor_date(
      sensing_date,
      "month"
    )
  )
```

```{r}
sensor_monthly <- covid_daily |>
  group_by(
    location_id,
    month_date,
    month
  ) |>
  summarise(
    mean_daily_count = mean(
      daily_count,
      na.rm = TRUE
    ),
    days_observed = n_distinct(sensing_date),
    .groups = "drop"
  ) |>
  filter(
    days_observed >= 20
  )
```

```{r}
pre_covid_baseline <- sensor_monthly |>
  filter(
    year(month_date) >= 2015,
    year(month_date) <= 2019
  ) |>
  group_by(
    location_id,
    month
  ) |>
  summarise(
    baseline_count = mean(
      mean_daily_count,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

```{r}
covid_index <- sensor_monthly |>
  left_join(
    pre_covid_baseline,
    by = c(
      "location_id",
      "month"
    )
  ) |>
  filter(
    !is.na(baseline_count),
    baseline_count > 0
  ) |>
  mutate(
    activity_index =
      mean_daily_count / baseline_count * 100
  )
```

```{r}
melbourne_covid_index <- covid_index |>
  group_by(month_date) |>
  summarise(
    activity_index = median(
      activity_index,
      na.rm = TRUE
    ),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

```{r}
melbourne_covid_index |>
  ggplot(
    aes(
      x = month_date,
      y = activity_index
    )
  ) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  geom_line(
    linewidth = 0.9
  ) +
  labs(
    title = "COVID-19 disrupted Melbourne's pedestrian pulse",
    subtitle = "Monthly activity relative to the same-month 2015–2019 baseline",
    x = NULL,
    y = "Pedestrian activity index\n(100 = pre-COVID baseline)"
  ) +
  theme_minimal()
```

```{r}
library(data.table)
library(ggplot2)
library(ggrepel)

proc <- fread("../pedestrian_counts_processed.csv")
raw <- fread("../Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv")

pre <- proc[year == 2019, .(daily = sum(pedestriancount)), by = .(location_id, sensing_date)]
pre <- pre[, .(pre = mean(daily)), by = location_id]

raw[, date := as.Date(paste(Year, Mdate, Month), format = "%Y %d %B")]
post <- raw[Year == 2022, .(daily = sum(Hourly_Counts)), by = .(Sensor_ID, date)]
post <- post[, .(post = mean(daily)), by = Sensor_ID]

chg <- merge(pre, post, by.x = "location_id", by.y = "Sensor_ID")
sname <- proc[, .(sensor_name = sensor_name[1]), by = location_id]
chg <- merge(chg, sname, by = "location_id")
chg[, pct := (post / pre - 1) * 100]
chg[, direction := fifelse(pct >= 0, "up", "down")]
chg[, sensor_name := reorder(sensor_name, pct)]

p5 <- ggplot(chg, aes(x = pct, y = sensor_name)) +
  geom_segment(aes(x = 0, xend = pct, yend = sensor_name, colour = direction)) +
  geom_point(aes(colour = direction), size = 2.2) +
  geom_text(aes(label = sprintf("%.0f%%", pct),
                hjust = ifelse(pct >= 0, -0.3, 1.3)), size = 2.6) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c(up = "#1a9850", down = "#d73027"),
                      labels = c(up = "Increased", down = "Decreased")) +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.15))) +
  labs(title = "Post-COVID change in daily pedestrian counts by sensor location",
       subtitle = "Mean daily counts, 2019 vs 2022, common sensor locations",
       x = "% change vs 2019 baseline", y = NULL, colour = NULL) +
  theme_minimal(base_size = 11)

ggsave("rq5_sensor_change.png", p5, width = 9, height = 11, dpi = 300)

daily <- proc[, .(total = sum(pedestriancount)), by = sensing_date]
daily[, date := as.Date(sensing_date)]
daily[, yr := as.integer(format(date, "%Y"))]
daily[, mo := as.integer(format(date, "%m"))]
daily[, dow := as.integer(format(date, "%u"))]

base <- daily[yr <= 2019, .(med = median(total)), by = .(mo, dow)]
daily <- merge(daily, base, by = c("mo", "dow"))
daily[, resid := total - med]

k <- 12
busy <- daily[order(-resid)][1:k, date]
quiet <- daily[order(resid)][1:k, date]
daily[, type := "normal"]
daily[date %in% busy, type := "busy"]
daily[date %in% quiet, type := "quiet"]

ev <- proc[, .(mega_event = mega_event[1],
               is_public_holiday = is_public_holiday[1],
               covid_phase = covid_phase[1]), by = .(date = as.Date(sensing_date))]
daily <- merge(daily, ev, by = "date", all.x = TRUE)

daily[, explain := fifelse(mega_event, "mega event",
        fifelse(is_public_holiday, "public holiday",
        fifelse(covid_phase != "Normal", covid_phase, "")))]
daily[type != "normal" & explain == "", explain := format(date, "%Y-%m-%d")]

covid_span <- daily[covid_phase != "Normal", range(date)]

p6 <- ggplot(daily, aes(date, resid)) +
  annotate("rect", xmin = covid_span[1], xmax = covid_span[2],
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35) +
  geom_line(colour = "grey70", linewidth = 0.4) +
  geom_point(data = daily[type != "normal"], aes(colour = type), size = 1.8) +
  geom_text_repel(data = daily[type != "normal"], aes(label = explain),
                  size = 2.6, max.overlaps = Inf, min.segment.length = 0,
                  segment.size = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(busy = "#d73027", quiet = "#2166ac"),
                      labels = c(busy = "Unusually busy", quiet = "Unusually quiet")) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(title = "Days unusually busy or quiet vs calendar-adjusted expectation",
       subtitle = "Calendar baseline: median count for the same month and day of week, pre-COVID years",
       x = NULL, y = "Daily count minus calendar expectation", colour = NULL) +
  theme_minimal(base_size = 11)

ggsave("rq6_outlier_days.png", p6, width = 12, height = 6.5, dpi = 300)

print(p5)
print(p6)
```


