---
title: "Pedestrian EDA"
format: html
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
```

```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )
```

```{r}

range(pedestrians$sensing_date, na.rm = TRUE)
```

```{r}

pedestrians %>%
  group_by(Year) %>%
  summarise(
    observations = n(),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

## Data Preparation

# Load historical pedestrian data

```{r}
historical_file <- "../Data_pedestrians/raw/Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv"

historical_raw <- read_csv(
  historical_file,
  show_col_types = FALSE
)

```

# Standardise historical data

```{r}

historical_clean <- historical_raw %>%
  mutate(
    sensing_date = mdy(
      str_c(Month, Mdate, Year, sep = " ")
    )
  ) %>%
  group_by(Sensor_ID, sensing_date) %>%
  arrange(ID, .by_group = TRUE) %>%
  mutate(
    hour_repaired = n() == 24 & n_distinct(Time) == 1,
    hourday = if_else(
      hour_repaired,
      as.integer(row_number() - 1),
      as.integer(Time)
    )
  ) %>%
  ungroup() %>%
  transmute(
    location_id = Sensor_ID,
    sensing_date,
    hourday,
    pedestrian_count = Hourly_Counts,
    source = "Historical"
  )
```


```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )

```


```{r}

glimpse(pedestrians)
```

# Standardise current API data

```{r}

current_clean <- pedestrians %>%
  transmute(
    location_id,
    sensing_date,
    hourday = as.integer(hourday),
    pedestrian_count = pedestriancount,
    source = "Current API"
  )
```

# Combine historical and current data

```{r}
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) %>%
  mutate(
    year = year(sensing_date)
  ) %>%
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

### Check Combined Data Coverage

```{r}
sort(unique(pedestrian_full$year))

```

## Daily and Weekly Pedestrian Rhythm

### Daily Rhythm Across the Available Years, 2009–2026


#### Calculate the average hourly pattern for each sensor within each year


```{r}

hourly_by_sensor_year <- pedestrian_full %>%
  filter(
    !is.na(hourday),
    hourday >= 0,
    hourday <= 23
  ) %>%
  group_by(year, location_id, hourday) %>%
  summarise(
    sensor_hour_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Average across sensors within each year

```{r}

hourly_by_year <- hourly_by_sensor_year %>%
  group_by(year, hourday) %>%
  summarise(
    avg_pedestrians = mean(sensor_hour_avg, na.rm = TRUE),
    .groups = "drop"
  )
hourly_index <- hourly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Typical Daily Pedestrian Rhythm

```{r}
ggplot(
  hourly_summary,
  aes(x = hourday, y = median_index)
) +
  geom_ribbon(
    aes(ymin = q25, ymax = q75),
    alpha = 0.2
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Typical Daily Pedestrian Rhythm Across Years",
    subtitle = "Median pattern across years; shaded area shows the middle 50% of years",
    x = "Hour of day",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()

```

#### Daily Rhythm by Year

```{r}
ggplot(
  hourly_index,
  aes(
    x = hourday,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_x_continuous(breaks = 0:23) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Daily Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Hour of day",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```

### Weekly Rhythm Across the Years, 2009–2026

```{r}
pedestrian_full <- pedestrian_full %>%
  mutate(
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )

```

#### Calculate Sensor-Level Weekly Patterns

```{r}
weekly_by_sensor_year <- pedestrian_full %>%
  group_by(year, location_id, day_name) %>%
  summarise(
    sensor_day_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Calculate Year-Level Weekly Patterns

```{r}

weekly_by_year <- weekly_by_sensor_year %>%
  group_by(year, day_name) %>%
  summarise(
    avg_pedestrians = mean(sensor_day_avg, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Standardise Weekly Activity by Year

```{r}
weekly_index <- weekly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
```

#### Summarise the Typical Weekly Pattern

```{r}

weekly_summary <- weekly_index %>%
  group_by(day_name) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    .groups = "drop"
  )

```

### Typical Weekly Pedestrian Rhythm

```{r}
ggplot(
  weekly_summary,
  aes(x = day_name, y = median_index, group = 1)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  labs(
    title = "Typical Weekly Pedestrian Rhythm Across Years",
    subtitle = "Median weekly pattern across years",
    x = "Day of week",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()
```

### Weekly Rhythm by Year

```{r}
ggplot(
  weekly_index,
  aes(
    x = day_name,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Weekly Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Day of week",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```



## Long-Term Change in Pedestrian Activity

### Check Year Coverage

```{r}
year_coverage <- pedestrian_full %>%
  group_by(year) %>%
  summarise(
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    sensors = n_distinct(location_id),
    observations = n(),
    .groups = "drop"
  )

print(year_coverage, n = Inf)

```

### Identify Full Calendar Years


```{r}
full_years <- year_coverage %>%
  filter(
    month(first_date) == 1,
    day(first_date) == 1,
    month(last_date) == 12,
    day(last_date) == 31
  ) %>%
  pull(year)

full_years
```

### Identify Common Sensors

```{r}
common_sensors <- pedestrian_full %>%
  filter(year %in% full_years) %>%
  distinct(year, location_id) %>%
  count(location_id, name = "years_present") %>%
  filter(years_present == length(full_years))

nrow(common_sensors)
```

### Calculate Annual Activity at Common Sensors

```{r}
common_sensor_ids <- common_sensors %>%
  pull(location_id)

common_sensor_year <- pedestrian_full %>%
  filter(
    year %in% full_years,
    location_id %in% common_sensor_ids
  ) %>%
  group_by(year, location_id) %>%
  summarise(
    avg_hourly_count = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

```

### Calculate Typical Activity for Each Year

```{r}
common_yearly_activity <- common_sensor_year %>%
  group_by(year) %>%
  summarise(
    typical_activity = median(avg_hourly_count, na.rm = TRUE),
    mean_activity = mean(avg_hourly_count, na.rm = TRUE),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

print(common_yearly_activity, n = Inf)

```


### Monthly Pedestrian Activity Over Time

```{r}
monthly_common_activity <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date, location_id) %>%
  summarise(
    sensor_month_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(month_date) %>%
  summarise(
    typical_activity = median(sensor_month_avg, na.rm = TRUE),
    .groups = "drop"
  )
```

##### Check Monthly Sensor Coverage

```{r}
monthly_sensor_coverage <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date) %>%
  summarise(
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

range(monthly_sensor_coverage$sensors)

monthly_sensor_coverage %>%
  count(sensors)

monthly_sensor_coverage %>%
  filter(sensors < 14)

```

#### Preserve Missing Periods

```{r}
monthly_common_activity <- monthly_common_activity %>%
  complete(
    month_date = seq.Date(
      min(month_date),
      max(month_date),
      by = "month"
    )
  )
```

### Long-Term Monthly Pedestrian Trend

```{r}
ggplot(
  monthly_common_activity,
  aes(x = month_date, y = typical_activity)
) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Long-Term Pedestrian Activity Across Available Data",
    subtitle = "Monthly median activity at common sensor locations",
    x = "Year",
    y = "Median hourly pedestrian count"
  ) +
  theme_minimal()

```



