---
title: "Pedestrian EDA"
format: html
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
```

```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )
```

```{r}

range(pedestrians$sensing_date, na.rm = TRUE)
```

```{r}

pedestrians %>%
  group_by(Year) %>%
  summarise(
    observations = n(),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

## Data Preparation

# Load historical pedestrian data

```{r}
historical_file <- "../Data_pedestrians/raw/Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv"

historical_raw <- read_csv(
  historical_file,
  show_col_types = FALSE
)

```

# Standardise historical data

```{r}

historical_clean <- historical_raw %>%
  mutate(
    sensing_date = mdy(
      str_c(Month, Mdate, Year, sep = " ")
    )
  ) %>%
  group_by(Sensor_ID, sensing_date) %>%
  arrange(ID, .by_group = TRUE) %>%
  mutate(
    hour_repaired = n() == 24 & n_distinct(Time) == 1,
    hourday = if_else(
      hour_repaired,
      as.integer(row_number() - 1),
      as.integer(Time)
    )
  ) %>%
  ungroup() %>%
  transmute(
    location_id = Sensor_ID,
    sensing_date,
    hourday,
    pedestrian_count = Hourly_Counts,
    source = "Historical"
  )
```


```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )

```


```{r}

glimpse(pedestrians)
```

# Standardise current API data

```{r}

current_clean <- pedestrians %>%
  transmute(
    location_id,
    sensing_date,
    hourday = as.integer(hourday),
    pedestrian_count = pedestriancount,
    source = "Current API"
  )
```

# Combine historical and current data

```{r}
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) %>%
  mutate(
    year = year(sensing_date)
  ) %>%
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

### Check Combined Data Coverage

```{r}
sort(unique(pedestrian_full$year))

```

## Daily and Weekly Pedestrian Rhythm

### Daily Rhythm Across the Available Years, 2009–2026


#### Calculate the average hourly pattern for each sensor within each year


```{r}

hourly_by_sensor_year <- pedestrian_full %>%
  filter(
    !is.na(hourday),
    hourday >= 0,
    hourday <= 23
  ) %>%
  group_by(year, location_id, hourday) %>%
  summarise(
    sensor_hour_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Average across sensors within each year

```{r}

hourly_by_year <- hourly_by_sensor_year %>%
  group_by(year, hourday) %>%
  summarise(
    avg_pedestrians = mean(sensor_hour_avg, na.rm = TRUE),
    .groups = "drop"
  )
hourly_index <- hourly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Typical Daily Pedestrian Rhythm

```{r}
ggplot(
  hourly_summary,
  aes(x = hourday, y = median_index)
) +
  geom_ribbon(
    aes(ymin = q25, ymax = q75),
    alpha = 0.2
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Typical Daily Pedestrian Rhythm Across Years",
    subtitle = "Median pattern across years; shaded area shows the middle 50% of years",
    x = "Hour of day",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()

```

#### Daily Rhythm by Year

```{r}
ggplot(
  hourly_index,
  aes(
    x = hourday,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_x_continuous(breaks = 0:23) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Daily Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Hour of day",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```

### Weekly Rhythm Across the Years, 2009–2026

```{r}
pedestrian_full <- pedestrian_full %>%
  mutate(
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )

```

#### Calculate Sensor-Level Weekly Patterns

```{r}
weekly_by_sensor_year <- pedestrian_full %>%
  group_by(year, location_id, day_name) %>%
  summarise(
    sensor_day_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Calculate Year-Level Weekly Patterns

```{r}

weekly_by_year <- weekly_by_sensor_year %>%
  group_by(year, day_name) %>%
  summarise(
    avg_pedestrians = mean(sensor_day_avg, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Standardise Weekly Activity by Year

```{r}
weekly_index <- weekly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
```

#### Summarise the Typical Weekly Pattern

```{r}

weekly_summary <- weekly_index %>%
  group_by(day_name) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    .groups = "drop"
  )

```

### Typical Weekly Pedestrian Rhythm

```{r}
ggplot(
  weekly_summary,
  aes(x = day_name, y = median_index, group = 1)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  labs(
    title = "Typical Weekly Pedestrian Rhythm Across Years",
    subtitle = "Median weekly pattern across years",
    x = "Day of week",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()
```

### Weekly Rhythm by Year

```{r}
ggplot(
  weekly_index,
  aes(
    x = day_name,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Weekly Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Day of week",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```



## Long-Term Change in Pedestrian Activity

### Check Year Coverage

```{r}
year_coverage <- pedestrian_full %>%
  group_by(year) %>%
  summarise(
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    sensors = n_distinct(location_id),
    observations = n(),
    .groups = "drop"
  )

print(year_coverage, n = Inf)

```

### Identify Full Calendar Years


```{r}
full_years <- year_coverage %>%
  filter(
    month(first_date) == 1,
    day(first_date) == 1,
    month(last_date) == 12,
    day(last_date) == 31
  ) %>%
  pull(year)

full_years
```

### Identify Common Sensors

```{r}
common_sensors <- pedestrian_full %>%
  filter(year %in% full_years) %>%
  distinct(year, location_id) %>%
  count(location_id, name = "years_present") %>%
  filter(years_present == length(full_years))

nrow(common_sensors)
```

### Calculate Annual Activity at Common Sensors

```{r}
common_sensor_ids <- common_sensors %>%
  pull(location_id)

common_sensor_year <- pedestrian_full %>%
  filter(
    year %in% full_years,
    location_id %in% common_sensor_ids
  ) %>%
  group_by(year, location_id) %>%
  summarise(
    avg_hourly_count = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

```

### Calculate Typical Activity for Each Year

```{r}
common_yearly_activity <- common_sensor_year %>%
  group_by(year) %>%
  summarise(
    typical_activity = median(avg_hourly_count, na.rm = TRUE),
    mean_activity = mean(avg_hourly_count, na.rm = TRUE),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

print(common_yearly_activity, n = Inf)

```


### Monthly Pedestrian Activity Over Time

```{r}
monthly_common_activity <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date, location_id) %>%
  summarise(
    sensor_month_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(month_date) %>%
  summarise(
    typical_activity = median(sensor_month_avg, na.rm = TRUE),
    .groups = "drop"
  )
```

##### Check Monthly Sensor Coverage

```{r}
monthly_sensor_coverage <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date) %>%
  summarise(
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

range(monthly_sensor_coverage$sensors)

monthly_sensor_coverage %>%
  count(sensors)

monthly_sensor_coverage %>%
  filter(sensors < 14)

```

#### Preserve Missing Periods

```{r}
monthly_common_activity <- monthly_common_activity %>%
  complete(
    month_date = seq.Date(
      min(month_date),
      max(month_date),
      by = "month"
    )
  )
```

### Long-Term Monthly Pedestrian Trend

```{r}
ggplot(
  monthly_common_activity,
  aes(x = month_date, y = typical_activity)
) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Long-Term Pedestrian Activity Across Available Data",
    subtitle = "Monthly median activity at common sensor locations",
    x = "Year",
    y = "Median hourly pedestrian count"
  ) +
  theme_minimal()

```


### Identify the Largest Drops in Pedestrian Activity


```{r}
### Identify the Largest Drops in Pedestrian Activity

month_quality <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date) %>%
  summarise(
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    days_observed = n_distinct(sensing_date),
    sensors = n_distinct(location_id),
    .groups = "drop"
  ) %>%
  mutate(
    full_month = days_observed == days_in_month(month_date)
  )
```


```{r}
monthly_changes <- monthly_common_activity %>%
  left_join(
    month_quality %>%
      select(month_date, sensors, full_month),
    by = "month_date"
  ) %>%
  arrange(month_date) %>%
  mutate(
    previous_month = lag(month_date),
    previous_activity = lag(typical_activity),
    previous_sensors = lag(sensors),
    previous_full_month = lag(full_month),

    change = typical_activity - previous_activity,

    pct_change =
      (typical_activity - previous_activity) /
      previous_activity * 100
  )
```


```{r}
biggest_dips <- monthly_changes %>%
  filter(
    full_month == TRUE,
    previous_full_month == TRUE,
    sensors >= 13,
    previous_sensors >= 13,
    !is.na(pct_change)
  ) %>%
  arrange(pct_change) %>%
  slice_head(n = 5) %>%
  select(
    previous_month,
    month_date,
    previous_activity,
    typical_activity,
    change,
    pct_change
  )

biggest_dips
```


```{r}
june_july_changes <- monthly_changes %>%
  filter(
    month(previous_month) == 6,
    month(month_date) == 7,
    !is.na(pct_change)
  ) %>%
  arrange(pct_change) %>%
  select(
    previous_month,
    month_date,
    pct_change
  )

june_july_changes

```

```{r}
sensor_change_2017 <- pedestrian_full %>%
  filter(
    location_id %in% common_sensor_ids,
    sensing_date >= as.Date("2017-06-01"),
    sensing_date <= as.Date("2017-07-31")
  ) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(location_id, month_date) %>%
  summarise(
    avg_activity = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = month_date,
    values_from = avg_activity
  ) %>%
  mutate(
    pct_change =
      (`2017-07-01` - `2017-06-01`) /
      `2017-06-01` * 100
  ) %>%
  arrange(pct_change)

sensor_change_2017

```



## EDA of City of Melbourne Snapshot Data

```{r}
library(tidyverse)
library(data.table)
library(lubridate)
```

### Locate Snapshot Files

```{r}
snapshot_dir <- here::here(
  "Data_pedestrians",
  "raw",
  "data_pedestrian"
)

snapshot_files <- list.files(
  snapshot_dir,
  pattern = "\\.csv$",
  full.names = TRUE
)

snapshot_files
```

### Inspect Snapshot File Structure

```{r}
snapshot_schema <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      nrows = 5
    )
    
    tibble::tibble(
      file = basename(f),
      n_columns = ncol(x),
      columns = paste(names(x), collapse = " | ")
    )
  }
)

snapshot_schema
```

### Check Snapshot Data Coverage

```{r}
snapshot_coverage <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c(
        "Location_ID",
        "Sensing_Date",
        "HourDay",
        "PedestrianCount"
      )
    )
    
    data.frame(
      file = basename(f),
      first_date = min(as.Date(x$Sensing_Date)),
      last_date = max(as.Date(x$Sensing_Date)),
      days = data.table::uniqueN(x$Sensing_Date),
      sensors = data.table::uniqueN(x$Location_ID),
      observations = nrow(x),
      hours = data.table::uniqueN(x$HourDay)
    )
  }
)

snapshot_coverage
```


### Check Daily Coverage


```{r}
snapshot_daily_coverage <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c("Sensing_Date")
    )
    
    dates <- as.Date(x$Sensing_Date)
    
    data.frame(
      file = basename(f),
      days_observed = data.table::uniqueN(dates),
      expected_days = ifelse(
        lubridate::leap_year(min(dates)),
        366,
        365
      ),
      missing_days = ifelse(
        lubridate::leap_year(min(dates)),
        366,
        365
      ) - data.table::uniqueN(dates)
    )
  }
)

snapshot_daily_coverage
```


### Check Duplicate Records

```{r}
snapshot_duplicates <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c(
        "Location_ID",
        "Sensing_Date",
        "HourDay"
      )
    )
    
    dup <- x[
      ,
      .N,
      by = .(Location_ID, Sensing_Date, HourDay)
    ][N > 1]
    
    data.frame(
      file = basename(f),
      duplicate_groups = nrow(dup),
      duplicated_rows = sum(dup$N - 1)
    )
  }
)

snapshot_duplicates
```


## Validation of Snapshot Data

### Validate the 2022 Snapshot

### Load 2022 Snapshot

```{r}
snapshot_2022 <- data.table::fread(
  file.path(snapshot_dir, "2022_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)

historical_2022 <- pedestrian_full %>%
  filter(year == 2022) %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count
  )
```


### Compare Snapshot with Existing Data

```{r}
comparison_2022 <- snapshot_2022 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    historical_2022,
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    count_difference = PedestrianCount - pedestrian_count
  )

summary(comparison_2022$count_difference)
```


### Summarise 2022 Differences


```{r}
comparison_2022 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(count_difference != 0, na.rm = TRUE),
    max_difference = max(abs(count_difference), na.rm = TRUE)
  )
```


### Examine Different Records

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  summarise(
    different_rows = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    mean_absolute_difference = mean(abs(count_difference)),
    median_absolute_difference = median(abs(count_difference)),
    max_absolute_difference = max(abs(count_difference))
  )
```


### Examine Differences by Month

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  mutate(
    month = lubridate::floor_date(Sensing_Date, "month")
  ) %>%
  count(month, sort = TRUE)
```


### Examine Differences by Sensor


```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  count(Location_ID, sort = TRUE)
```


### Inspect Largest Differences

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  arrange(desc(abs(count_difference))) %>%
  select(
    Location_ID,
    Sensing_Date,
    HourDay,
    PedestrianCount,
    pedestrian_count,
    count_difference
  ) %>%
  head(20)
```


### Investigate Location 6


```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  summarise(
    n = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    min_difference = min(count_difference),
    max_difference = max(count_difference)
  )
```


### Location 6 Differences by Month

```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  mutate(
    month = lubridate::floor_date(Sensing_Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    different_records = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    .groups = "drop"
  )
```

### Direction of Differences

```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  count(
    direction = case_when(
      count_difference > 0 ~ "Snapshot higher",
      count_difference < 0 ~ "Snapshot lower",
      TRUE ~ "Same"
    )
  )
```


### Validate Direction Totals


```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  arrange(desc(abs(count_difference))) %>%
  select(
    Location_ID,
    Sensing_Date,
    HourDay,
    PedestrianCount,
    pedestrian_count,
    count_difference
  ) %>%
  head(20)
```

```{r}
snapshot_2022_full <- data.table::fread(
  file.path(snapshot_dir, "2022_Pedestrian_Data.csv")
)

snapshot_2022_full %>%
  mutate(
    direction_total = Direction_1 + Direction_2,
    difference = PedestrianCount - direction_total
  ) %>%
  summarise(
    rows = n(),
    exact_match = sum(difference == 0, na.rm = TRUE),
    different = sum(difference != 0, na.rm = TRUE),
    max_difference = max(abs(difference), na.rm = TRUE)
  )
```

```{r}
snapshot_2022_full %>%
  filter(Location_ID == 6) %>%
  summarise(
    mean_pedestrian_count = mean(PedestrianCount, na.rm = TRUE),
    mean_direction_total = mean(Direction_1 + Direction_2, na.rm = TRUE),
    mean_difference = mean(
      PedestrianCount - (Direction_1 + Direction_2),
      na.rm = TRUE
    )
  )
```

## Validate the 2024 Snapshot
### Load 2024 Snapshot

```{r}
snapshot_2024 <- data.table::fread(
  file.path(snapshot_dir, "2024_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)
```

### Compare with Existing 2024 Data

```{r}
comparison_2024 <- snapshot_2024 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    pedestrian_full %>%
      filter(lubridate::year(sensing_date) == 2024) %>%
      select(
        location_id,
        sensing_date,
        hourday,
        pedestrian_count
      ),
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    difference = PedestrianCount - pedestrian_count
  )

comparison_2024 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(difference != 0, na.rm = TRUE),
    mean_difference = mean(difference, na.rm = TRUE),
    max_absolute_difference = max(abs(difference), na.rm = TRUE)
  )
```

## Validate the 2025 Snapshot

### Load 2025 Snapshot

```{r}

snapshot_2025 <- data.table::fread(
  file.path(snapshot_dir, "2025_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)
```


```{r}
comparison_2025 <- snapshot_2025 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    pedestrian_full %>%
      filter(lubridate::year(sensing_date) == 2025) %>%
      select(
        location_id,
        sensing_date,
        hourday,
        pedestrian_count
      ),
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    difference = PedestrianCount - pedestrian_count
  )


comparison_2025 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(difference != 0, na.rm = TRUE),
    mean_difference = mean(difference, na.rm = TRUE),
    max_absolute_difference = max(abs(difference), na.rm = TRUE)
  )
```


## Standardisation and Integration

### Standardise Snapshot Data

```{r}
snapshot_standardised <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    
    x <- data.table::fread(file = f) %>%
      as_tibble()
    
    x %>%
      transmute(
        location_id = as.numeric(Location_ID),
        sensing_date = as.Date(Sensing_Date),
        hourday = as.numeric(HourDay),
        pedestrian_count = as.numeric(PedestrianCount),
        hour_repaired = FALSE,
        source = paste0("city_snapshot_", substr(basename(f), 1, 4))
      )
  }
) %>%
  mutate(
    year = lubridate::year(sensing_date),
    month = lubridate::month(sensing_date),
    month_name = lubridate::month(sensing_date, label = TRUE),
    day_name = lubridate::wday(sensing_date, label = TRUE),
    day_of_week = lubridate::wday(sensing_date, week_start = 1),
    is_weekend = day_of_week >= 6,
    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  )
```


### Check Standardised Snapshot Data


```{r}
snapshot_standardised %>%
  count(year)
```

```{r}
snapshot_standardised %>%
  summarise(
    rows = n(),
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    sensors = n_distinct(location_id),
    duplicate_groups = n() -
      n_distinct(location_id, sensing_date, hourday)
  )
```

## Replace Existing Data

### Replace 2022–2025 with Validated Snapshots

```{r}


## Rebuild final pedestrian dataset

### Keep historical data before 2022

historical_data <- pedestrian_full %>%
  filter(!year %in% c(2022, 2023, 2024, 2025)) %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count,
    source,
    any_of("hour_repaired")
  )

# If the historical dataset does not contain hour_repaired,
# create it as FALSE.
if (!"hour_repaired" %in% names(historical_data)) {
  historical_data$hour_repaired <- FALSE
}


### Prepare validated 2022–2025 snapshots

snapshot_data <- snapshot_standardised %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count,
    source
  ) %>%
  mutate(
    hour_repaired = FALSE
  )


### Combine historical and validated data

pedestrian_full_updated <- bind_rows(
  historical_data,
  snapshot_data
) %>%
  mutate(
    year = lubridate::year(sensing_date),

    month = lubridate::month(sensing_date),

    month_name = lubridate::month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),

    day_name = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),

    day_of_week = lubridate::wday(
      sensing_date,
      week_start = 1
    ),

    is_weekend = day_of_week >= 6,

    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  ) %>%
  arrange(
    sensing_date,
    hourday,
    location_id
)

```

### Recreate Derived Variables

```{r}
pedestrian_full_updated %>%
  group_by(year) %>%
  summarise(
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    observations = n(),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

## Final Validation

### Check Yearly Coverage

```{r}
final_check <- pedestrian_full_updated %>%
  summarise(
    total_rows = n(),
    unique_sensor_hour_records =
      n_distinct(location_id, sensing_date, hourday),
    duplicate_rows =
      n() - n_distinct(location_id, sensing_date, hourday),
    missing_counts =
      sum(is.na(pedestrian_count)),
    first_date = min(sensing_date),
    last_date = max(sensing_date)
  )

final_check
```

```{r}
pedestrian_full_updated %>%
  count(year)
```

```{r}
pedestrian_full <- pedestrian_full_updated
```


## Daily and Weekly Pedestrian Rhythm

### Hourly Pedestrian Activity Over Time

```{r}
library(plotly)
```

```{r}
hourly_by_sensor_year <- pedestrian_full %>%
  filter(
    !is.na(hourday),
    hourday >= 0,
    hourday <= 23
  ) %>%
  group_by(year, location_id, hourday) %>%
  summarise(
    sensor_hour_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

hourly_by_year <- hourly_by_sensor_year %>%
  group_by(year, hourday) %>%
  summarise(
    avg_pedestrians = mean(sensor_hour_avg, na.rm = TRUE),
    .groups = "drop"
  )

hourly_index <- hourly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()

hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
daily_interactive_data <- hourly_index %>%
  mutate(
    year_label = case_when(
      year == 2009 ~ "2009 (partial)",
      year == 2026 ~ "2026 (partial)",
      TRUE ~ as.character(year)
    )
  )
```

### PLOT 1 — Typical Daily Pedestrian Rhythm

```{r}

daily_interactive <- plot_ly(
  daily_interactive_data,
  x = ~hourday,
  y = ~activity_index,
  color = ~year_label,
  type = "scatter",
  mode = "lines",
  text = ~paste0(
    "Year: ", year_label,
    "<br>Hour: ", hourday, ":00",
    "<br>Activity index: ", round(activity_index, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Daily Pedestrian Rhythm by Year",
    xaxis = list(
      title = "Hour of day",
      tickmode = "linear",
      tick0 = 0,
      dtick = 1
    ),
    yaxis = list(
      title = "Activity index (100 = yearly average)"
    ),
    hovermode = "closest",
    legend = list(
      title = list(text = "Year"),
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    margin = list(r = 150)
  )

daily_interactive
```

## Weekly Rhythm Across the Years, 2009–2026

```{r}

pedestrian_full <- pedestrian_full %>%
  mutate(
    day_name = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )
```

### Calculating Weekly Patterns

```{r}
weekly_by_sensor_year <- pedestrian_full %>%
  group_by(year, location_id, day_name) %>%
  summarise(
    sensor_day_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

weekly_by_year <- weekly_by_sensor_year %>%
  group_by(year, day_name) %>%
  summarise(
    avg_pedestrians = mean(sensor_day_avg, na.rm = TRUE),
    .groups = "drop"
  )

weekly_index <- weekly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()

weekly_summary <- weekly_index %>%
  group_by(day_name) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
weekly_interactive_data <- weekly_index %>%
  mutate(
    year_label = case_when(
      year == 2009 ~ "2009 (partial)",
      year == 2026 ~ "2026 (partial)",
      TRUE ~ as.character(year)
    )
  )
```

### PLOT 2 — Typical Weekly Pedestrian Rhythm

```{r}
weekly_interactive <- plot_ly(
  weekly_interactive_data,
  x = ~day_name,
  y = ~activity_index,
  color = ~year_label,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    "Year: ", year_label,
    "<br>Day: ", day_name,
    "<br>Activity index: ", round(activity_index, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Weekly Pedestrian Rhythm by Year",
    xaxis = list(
      title = "Day of week"
    ),
    yaxis = list(
      title = "Activity index (100 = yearly average)"
    ),
    hovermode = "closest",
    legend = list(
      title = list(text = "Year"),
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    margin = list(r = 150)
  )

weekly_interactive
```


### ### Daily pedestrian activity for dip analysis

```{r}

decrease_sensor_daily <- pedestrian_full %>%
  filter(
    !is.na(pedestrian_count)
  ) %>%
  group_by(
    location_id,
    sensing_date
  ) %>%
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    hours_observed = n_distinct(hourday),
    .groups = "drop"
  ) %>%
  filter(
    hours_observed >= 20
  )
```

### Calculate expected activity for each sensor and weekday

```{r}


decrease_sensor_index <- decrease_sensor_daily %>%
  arrange(
    location_id,
    sensing_date
  ) %>%
  mutate(
    day_of_week = lubridate::wday(
      sensing_date,
      week_start = 1
    )
  ) %>%
  group_by(
    location_id,
    day_of_week
  ) %>%
  mutate(
    expected_count = slider::slide_dbl(
      daily_count,
      ~ median(
        .x[-5],
        na.rm = TRUE
      ),
      .before = 4,
      .after = 4,
      .complete = TRUE
    )
  ) %>%
  ungroup() %>%
  mutate(
    decrease_percent =
      100 *
      (daily_count - expected_count) /
      expected_count
  ) %>%
  filter(
    is.finite(decrease_percent)
  )
```


### Summarise pedestrian decreases across sensors


```{r}


decrease_day_summary <- decrease_sensor_index %>%
  group_by(
    sensing_date
  ) %>%
  summarise(
    median_decrease = median(
      decrease_percent,
      na.rm = TRUE
    ),

    sensors_reporting = n_distinct(
      location_id
    ),

    sensors_below_20 = sum(
      decrease_percent <= -20,
      na.rm = TRUE
    ),

    proportion_below_20 = mean(
      decrease_percent <= -20,
      na.rm = TRUE
    ),

    median_daily_count = median(
      daily_count,
      na.rm = TRUE
    ),

    .groups = "drop"
  )
```

### Identify significant pedestrian decreases

```{r}

decrease_candidates <- decrease_day_summary %>%
  filter(
    sensors_reporting >= 10,
    median_decrease <= -20,
    proportion_below_20 >= 0.30
  ) %>%
  arrange(
    median_decrease
  )

decrease_candidates %>%
  slice_head(n = 30)
```

### Focus on non-COVID significant decreases

```{r}

decrease_non_covid <- decrease_candidates %>%
  filter(
    lubridate::year(sensing_date) < 2020 |
      lubridate::year(sensing_date) > 2022
  )
```


```{r}
decrease_non_covid %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20
  ) %>%
  slice_head(n = 30)
```

```{r}
decrease_non_covid %>%
  mutate(
    year = lubridate::year(sensing_date)
  ) %>%
  count(year)
```



```{r}
### Top non-COVID pedestrian decreases

top_decreases <- decrease_non_covid %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20
  ) %>%
  slice_head(n = 20) %>%
  mutate(
    day = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    )
  )

top_decreases
```



```{r}
top_decreases %>%
  mutate(
    month = lubridate::month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    )
  ) %>%
  count(month, sort = TRUE)
```

```{r}
top_decreases %>%
  mutate(
    month_day = format(
      sensing_date,
      "%d %B"
    )
  ) %>%
  count(month_day, sort = TRUE)
```


```{r}
top_decreases %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20,
    day
  ) %>%
  arrange(sensing_date)
```


```{r}
### Victorian public holidays, 2009–2026

easter_sunday <- function(y) {
  a <- y %% 19
  b <- y %/% 100
  c <- y %% 100
  d <- b %/% 4
  e <- b %% 4
  f <- (b + 8) %/% 25
  g <- (b - f + 1) %/% 3
  h <- (19 * a + b - d - g + 15) %% 30
  i <- c %/% 4
  k <- c %% 4
  l <- (32 + 2 * e + 2 * i - h - k) %% 7
  m <- (a + 11 * h + 22 * l) %/% 451

  month <- (h + l - 7 * m + 114) %/% 31
  day <- ((h + l - 7 * m + 114) %% 31) + 1

  as.Date(
    sprintf("%04d-%02d-%02d", y, month, day)
  )
}
```


```{r}
holiday_calendar <- purrr::map_dfr(
  2009:2026,
  function(y) {

    easter <- easter_sunday(y)

    first_march <- as.Date(
      paste0(y, "-03-01")
    )

    first_june <- as.Date(
      paste0(y, "-06-01")
    )

    first_november <- as.Date(
      paste0(y, "-11-01")
    )

    second_monday_march <-
      first_march +
      lubridate::days(
        (8 - lubridate::wday(
          first_march,
          week_start = 1
        )) %% 7 + 7
      )

    second_monday_june <-
      first_june +
      lubridate::days(
        (8 - lubridate::wday(
          first_june,
          week_start = 1
        )) %% 7 + 7
      )

    first_tuesday_november <-
      first_november +
      lubridate::days(
        (2 - lubridate::wday(
          first_november,
          week_start = 1
        )) %% 7
      )

    new_year <- as.Date(
      paste0(y, "-01-01")
    )

    australia_day <- as.Date(
      paste0(y, "-01-26")
    )

    anzac_day <- as.Date(
      paste0(y, "-04-25")
    )

    christmas <- as.Date(
      paste0(y, "-12-25")
    )

    boxing_day <- as.Date(
      paste0(y, "-12-26")
    )

    holidays <- tibble::tibble(
      holiday_date = c(
        new_year,
        australia_day,
        second_monday_march,
        easter - 2,
        easter - 1,
        easter,
        easter + 1,
        anzac_day,
        second_monday_june,
        first_tuesday_november,
        christmas,
        boxing_day
      ),

      holiday = c(
        "New Year's Day",
        "Australia Day",
        "Labour Day",
        "Good Friday",
        "Easter Saturday",
        "Easter Sunday",
        "Easter Monday",
        "ANZAC Day",
        if (y <= 2022) {
          "Queen's Birthday"
        } else {
          "King's Birthday"
        },
        "Melbourne Cup Day",
        "Christmas Day",
        "Boxing Day"
      )
    )

    # Substitute holidays
    if (lubridate::wday(
      new_year,
      week_start = 1
    ) %in% c(6, 7)) {

      holidays <- dplyr::bind_rows(
        holidays,
        tibble::tibble(
          holiday_date =
            new_year +
            lubridate::days(
              8 - lubridate::wday(
                new_year,
                week_start = 1
              )
            ),
          holiday = "New Year's Day (substitute)"
        )
      )
    }

    if (lubridate::wday(
      australia_day,
      week_start = 1
    ) %in% c(6, 7)) {

      holidays <- dplyr::bind_rows(
        holidays,
        tibble::tibble(
          holiday_date =
            australia_day +
            lubridate::days(
              8 - lubridate::wday(
                australia_day,
                week_start = 1
              )
            ),
          holiday = "Australia Day (substitute)"
        )
      )
    }

    if (lubridate::wday(
      christmas,
      week_start = 1
    ) %in% c(6, 7)) {

      holidays <- dplyr::bind_rows(
        holidays,
        tibble::tibble(
          holiday_date =
            as.Date(paste0(y, "-12-27")),
          holiday = "Christmas Day (substitute)"
        )
      )
    }

    if (lubridate::wday(
      boxing_day,
      week_start = 1
    ) %in% c(6, 7)) {

      holidays <- dplyr::bind_rows(
        holidays,
        tibble::tibble(
          holiday_date =
            as.Date(paste0(y, "-12-28")),
          holiday = "Boxing Day (substitute)"
        )
      )
    }

    holidays
  }
)
```


```{r}
top_decreases_holidays <- top_decreases %>%
  left_join(
    holiday_calendar,
    by = c(
      "sensing_date" = "holiday_date"
    )
  ) %>%
  mutate(
    holiday_status = if_else(
      is.na(holiday),
      "Normal day",
      "Public holiday"
    )
  )
```

```{r}
top_decreases_holidays %>%
  count(
    holiday_status
  )
```


```{r}
### Public holiday representation among all significant decreases

decrease_holiday_test <- decrease_non_covid %>%
  left_join(
    holiday_calendar,
    by = c("sensing_date" = "holiday_date")
  ) %>%
  mutate(
    holiday_status = if_else(
      is.na(holiday),
      "Normal day",
      "Public holiday"
    )
  )

decrease_holiday_test %>%
  count(holiday_status) %>%
  mutate(
    percentage = n / sum(n) * 100
  )
```


```{r}
decrease_holiday_test %>%
  filter(
    holiday_status == "Public holiday"
  ) %>%
  count(
    holiday,
    sort = TRUE
  )
```



```{r}
decrease_holiday_test %>%
  group_by(
    holiday_status
  ) %>%
  summarise(
    n = n(),
    median_decrease = median(
      median_decrease,
      na.rm = TRUE
    ),
    mean_decrease = mean(
      median_decrease,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```



```{r}
### Compare public holiday frequency with all eligible days

eligible_days <- decrease_day_summary %>%
  filter(
    sensors_reporting >= 10,
    lubridate::year(sensing_date) < 2020 |
      lubridate::year(sensing_date) > 2022
  ) %>%
  select(sensing_date) %>%
  distinct() %>%
  mutate(
    holiday_status = if_else(
      sensing_date %in% holiday_calendar$holiday_date,
      "Public holiday",
      "Normal day"
    )
  )

baseline_holiday <- eligible_days %>%
  count(holiday_status) %>%
  mutate(
    percentage = n / sum(n) * 100
  )

baseline_holiday
```


```{r}
### Compare significant dips with the overall calendar

significant_holiday <- decrease_holiday_test %>%
  count(holiday_status) %>%
  mutate(
    percentage = n / sum(n) * 100
  )

significant_holiday
```


```{r}
### Public holiday enrichment among significant decreases

holiday_enrichment <- tibble::tibble(
  baseline_holiday_percentage =
    baseline_holiday %>%
    filter(
      holiday_status == "Public holiday"
    ) %>%
    pull(percentage),

  significant_dip_holiday_percentage =
    significant_holiday %>%
    filter(
      holiday_status == "Public holiday"
    ) %>%
    pull(percentage)
) %>%
  mutate(
    enrichment =
      significant_dip_holiday_percentage /
      baseline_holiday_percentage
  )

holiday_enrichment
```




