---
title: "Pedestrian EDA"
format: html
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
```

```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )
```

```{r}

range(pedestrians$sensing_date, na.rm = TRUE)
```

```{r}

pedestrians %>%
  group_by(Year) %>%
  summarise(
    observations = n(),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

## Data Preparation


# Load historical pedestrian data

```{r}
historical_file <- "../Data_pedestrians/raw/Pedestrian_Counting_System_Monthly_counts_per_hour_may_2009_to_14_dec_2022.csv"

historical_raw <- read_csv(
  historical_file,
  show_col_types = FALSE
)

```

# Standardise historical data

```{r}

historical_clean <- historical_raw %>%
  mutate(
    sensing_date = mdy(
      str_c(Month, Mdate, Year, sep = " ")
    )
  ) %>%
  group_by(Sensor_ID, sensing_date) %>%
  arrange(ID, .by_group = TRUE) %>%
  mutate(
    hour_repaired = n() == 24 & n_distinct(Time) == 1,
    hourday = if_else(
      hour_repaired,
      as.integer(row_number() - 1),
      as.integer(Time)
    )
  ) %>%
  ungroup() %>%
  transmute(
    location_id = Sensor_ID,
    sensing_date,
    hourday,
    pedestrian_count = Hourly_Counts,
    source = "Historical"
  )
```


```{r}
api_file <- "../Data_pedestrians/raw2/pedestrian_counts_raw.csv"

pedestrians <- read_csv(
  api_file,
  col_select = c(
    sensing_date,
    hourday,
    location_id,
    sensor_name,
    pedestriancount,
    location
  ),
  show_col_types = FALSE
) %>%
  mutate(
    sensing_date = as.Date(sensing_date),
    Year = year(sensing_date)
  )

```


```{r}

glimpse(pedestrians)
```

# Standardise current API data

```{r}

current_clean <- pedestrians %>%
  transmute(
    location_id,
    sensing_date,
    hourday = as.integer(hourday),
    pedestrian_count = pedestriancount,
    source = "Current API"
  )
```

# Combine historical and current data

```{r}
pedestrian_full <- bind_rows(
  historical_clean,
  current_clean
) %>%
  mutate(
    year = year(sensing_date)
  ) %>%
  arrange(
    sensing_date,
    location_id,
    hourday
  )
```

### Check Combined Data Coverage

```{r}
sort(unique(pedestrian_full$year))

```

## Daily and Weekly Pedestrian Rhythm

### Daily Rhythm Across the Available Years, 2009–2026


#### Calculate the average hourly pattern for each sensor within each year


```{r}

hourly_by_sensor_year <- pedestrian_full %>%
  filter(
    !is.na(hourday),
    hourday >= 0,
    hourday <= 23
  ) %>%
  group_by(year, location_id, hourday) %>%
  summarise(
    sensor_hour_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Average across sensors within each year

```{r}

hourly_by_year <- hourly_by_sensor_year %>%
  group_by(year, hourday) %>%
  summarise(
    avg_pedestrians = mean(sensor_hour_avg, na.rm = TRUE),
    .groups = "drop"
  )
hourly_index <- hourly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Typical Daily Pedestrian Rhythm

```{r}
ggplot(
  hourly_summary,
  aes(x = hourday, y = median_index)
) +
  geom_ribbon(
    aes(ymin = q25, ymax = q75),
    alpha = 0.2
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Typical Daily Pedestrian Rhythm Across Years",
    subtitle = "Median pattern across years; shaded area shows the middle 50% of years",
    x = "Hour of day",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()

```

#### Daily Rhythm by Year

```{r}
ggplot(
  hourly_index,
  aes(
    x = hourday,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_x_continuous(breaks = 0:23) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Daily Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Hour of day",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```

### Weekly Rhythm Across the Years, 2009–2026

```{r}
pedestrian_full <- pedestrian_full %>%
  mutate(
    day_name = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )

```

#### Calculate Sensor-Level Weekly Patterns

```{r}
weekly_by_sensor_year <- pedestrian_full %>%
  group_by(year, location_id, day_name) %>%
  summarise(
    sensor_day_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Calculate Year-Level Weekly Patterns

```{r}

weekly_by_year <- weekly_by_sensor_year %>%
  group_by(year, day_name) %>%
  summarise(
    avg_pedestrians = mean(sensor_day_avg, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Standardise Weekly Activity by Year

```{r}
weekly_index <- weekly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()
```

#### Summarise the Typical Weekly Pattern

```{r}

weekly_summary <- weekly_index %>%
  group_by(day_name) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    .groups = "drop"
  )

```

### Typical Weekly Pedestrian Rhythm

```{r}
ggplot(
  weekly_summary,
  aes(x = day_name, y = median_index, group = 1)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed"
  ) +
  labs(
    title = "Typical Weekly Pedestrian Rhythm Across Years",
    subtitle = "Median weekly pattern across years",
    x = "Day of week",
    y = "Activity index (100 = yearly average)"
  ) +
  theme_minimal()
```

### Weekly Rhythm by Year

```{r}
ggplot(
  weekly_index,
  aes(
    x = day_name,
    y = factor(year, levels = 2009:2026),
    fill = activity_index
  )
) +
  geom_tile() +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "Melbourne's Weekly Pedestrian Rhythm Across Years",
    subtitle = "Activity relative to each year's average (100 = yearly average)",
    x = "Day of week",
    y = "Year",
    fill = "Activity\nindex"
  ) +
  theme_minimal()

```



## Long-Term Change in Pedestrian Activity

### Check Year Coverage

```{r}
year_coverage <- pedestrian_full %>%
  group_by(year) %>%
  summarise(
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    sensors = n_distinct(location_id),
    observations = n(),
    .groups = "drop"
  )

print(year_coverage, n = Inf)

```

### Identify Full Calendar Years


```{r}
full_years <- year_coverage %>%
  filter(
    month(first_date) == 1,
    day(first_date) == 1,
    month(last_date) == 12,
    day(last_date) == 31
  ) %>%
  pull(year)

full_years
```

### Identify Common Sensors

```{r}
common_sensors <- pedestrian_full %>%
  filter(year %in% full_years) %>%
  distinct(year, location_id) %>%
  count(location_id, name = "years_present") %>%
  filter(years_present == length(full_years))

nrow(common_sensors)
```

### Calculate Annual Activity at Common Sensors

```{r}
common_sensor_ids <- common_sensors %>%
  pull(location_id)

common_sensor_year <- pedestrian_full %>%
  filter(
    year %in% full_years,
    location_id %in% common_sensor_ids
  ) %>%
  group_by(year, location_id) %>%
  summarise(
    avg_hourly_count = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

```

### Calculate Typical Activity for Each Year

```{r}
common_yearly_activity <- common_sensor_year %>%
  group_by(year) %>%
  summarise(
    typical_activity = median(avg_hourly_count, na.rm = TRUE),
    mean_activity = mean(avg_hourly_count, na.rm = TRUE),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

print(common_yearly_activity, n = Inf)

```


### Monthly Pedestrian Activity Over Time

```{r}
monthly_common_activity <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date, location_id) %>%
  summarise(
    sensor_month_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(month_date) %>%
  summarise(
    typical_activity = median(sensor_month_avg, na.rm = TRUE),
    .groups = "drop"
  )
```

##### Check Monthly Sensor Coverage

```{r}
monthly_sensor_coverage <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date) %>%
  summarise(
    sensors = n_distinct(location_id),
    .groups = "drop"
  )

range(monthly_sensor_coverage$sensors)

monthly_sensor_coverage %>%
  count(sensors)

monthly_sensor_coverage %>%
  filter(sensors < 14)

```

#### Preserve Missing Periods

```{r}
monthly_common_activity <- monthly_common_activity %>%
  complete(
    month_date = seq.Date(
      min(month_date),
      max(month_date),
      by = "month"
    )
  )
```

### Long-Term Monthly Pedestrian Trend

```{r}
ggplot(
  monthly_common_activity,
  aes(x = month_date, y = typical_activity)
) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Long-Term Pedestrian Activity Across Available Data",
    subtitle = "Monthly median activity at common sensor locations",
    x = "Year",
    y = "Median hourly pedestrian count"
  ) +
  theme_minimal()

```


### Identify the Largest Drops in Pedestrian Activity


```{r}
### Identify the Largest Drops in Pedestrian Activity

month_quality <- pedestrian_full %>%
  filter(location_id %in% common_sensor_ids) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(month_date) %>%
  summarise(
    first_date = min(sensing_date, na.rm = TRUE),
    last_date = max(sensing_date, na.rm = TRUE),
    days_observed = n_distinct(sensing_date),
    sensors = n_distinct(location_id),
    .groups = "drop"
  ) %>%
  mutate(
    full_month = days_observed == days_in_month(month_date)
  )
```


```{r}
monthly_changes <- monthly_common_activity %>%
  left_join(
    month_quality %>%
      select(month_date, sensors, full_month),
    by = "month_date"
  ) %>%
  arrange(month_date) %>%
  mutate(
    previous_month = lag(month_date),
    previous_activity = lag(typical_activity),
    previous_sensors = lag(sensors),
    previous_full_month = lag(full_month),

    change = typical_activity - previous_activity,

    pct_change =
      (typical_activity - previous_activity) /
      previous_activity * 100
  )
```


```{r}
biggest_dips <- monthly_changes %>%
  filter(
    full_month == TRUE,
    previous_full_month == TRUE,
    sensors >= 13,
    previous_sensors >= 13,
    !is.na(pct_change)
  ) %>%
  arrange(pct_change) %>%
  slice_head(n = 5) %>%
  select(
    previous_month,
    month_date,
    previous_activity,
    typical_activity,
    change,
    pct_change
  )

biggest_dips
```


```{r}
june_july_changes <- monthly_changes %>%
  filter(
    month(previous_month) == 6,
    month(month_date) == 7,
    !is.na(pct_change)
  ) %>%
  arrange(pct_change) %>%
  select(
    previous_month,
    month_date,
    pct_change
  )

june_july_changes

```

```{r}
sensor_change_2017 <- pedestrian_full %>%
  filter(
    location_id %in% common_sensor_ids,
    sensing_date >= as.Date("2017-06-01"),
    sensing_date <= as.Date("2017-07-31")
  ) %>%
  mutate(
    month_date = floor_date(sensing_date, "month")
  ) %>%
  group_by(location_id, month_date) %>%
  summarise(
    avg_activity = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = month_date,
    values_from = avg_activity
  ) %>%
  mutate(
    pct_change =
      (`2017-07-01` - `2017-06-01`) /
      `2017-06-01` * 100
  ) %>%
  arrange(pct_change)

sensor_change_2017

```



## EDA of City of Melbourne Snapshot Data

```{r}
library(tidyverse)
library(data.table)
library(lubridate)
```

### Locate Snapshot Files

```{r}
snapshot_dir <- here::here(
  "Data_pedestrians",
  "raw",
  "data_pedestrian"
)

snapshot_files <- list.files(
  snapshot_dir,
  pattern = "\\.csv$",
  full.names = TRUE
)

snapshot_files
```

### Inspect Snapshot File Structure

```{r}
snapshot_schema <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      nrows = 5
    )
    
    tibble::tibble(
      file = basename(f),
      n_columns = ncol(x),
      columns = paste(names(x), collapse = " | ")
    )
  }
)

snapshot_schema
```

### Check Snapshot Data Coverage

```{r}
snapshot_coverage <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c(
        "Location_ID",
        "Sensing_Date",
        "HourDay",
        "PedestrianCount"
      )
    )
    
    data.frame(
      file = basename(f),
      first_date = min(as.Date(x$Sensing_Date)),
      last_date = max(as.Date(x$Sensing_Date)),
      days = data.table::uniqueN(x$Sensing_Date),
      sensors = data.table::uniqueN(x$Location_ID),
      observations = nrow(x),
      hours = data.table::uniqueN(x$HourDay)
    )
  }
)

snapshot_coverage
```


### Check Daily Coverage


```{r}
snapshot_daily_coverage <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c("Sensing_Date")
    )
    
    dates <- as.Date(x$Sensing_Date)
    
    data.frame(
      file = basename(f),
      days_observed = data.table::uniqueN(dates),
      expected_days = ifelse(
        lubridate::leap_year(min(dates)),
        366,
        365
      ),
      missing_days = ifelse(
        lubridate::leap_year(min(dates)),
        366,
        365
      ) - data.table::uniqueN(dates)
    )
  }
)

snapshot_daily_coverage
```


### Check Duplicate Records

```{r}
snapshot_duplicates <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    x <- data.table::fread(
      file = f,
      select = c(
        "Location_ID",
        "Sensing_Date",
        "HourDay"
      )
    )
    
    dup <- x[
      ,
      .N,
      by = .(Location_ID, Sensing_Date, HourDay)
    ][N > 1]
    
    data.frame(
      file = basename(f),
      duplicate_groups = nrow(dup),
      duplicated_rows = sum(dup$N - 1)
    )
  }
)

snapshot_duplicates
```


## Validation of Snapshot Data

### Validate the 2022 Snapshot

### Load 2022 Snapshot

```{r}
snapshot_2022 <- data.table::fread(
  file.path(snapshot_dir, "2022_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)

historical_2022 <- pedestrian_full %>%
  filter(year == 2022) %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count
  )
```


### Compare Snapshot with Existing Data

```{r}
comparison_2022 <- snapshot_2022 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    historical_2022,
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    count_difference = PedestrianCount - pedestrian_count
  )

summary(comparison_2022$count_difference)
```


### Summarise 2022 Differences


```{r}
comparison_2022 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(count_difference != 0, na.rm = TRUE),
    max_difference = max(abs(count_difference), na.rm = TRUE)
  )
```


### Examine Different Records

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  summarise(
    different_rows = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    mean_absolute_difference = mean(abs(count_difference)),
    median_absolute_difference = median(abs(count_difference)),
    max_absolute_difference = max(abs(count_difference))
  )
```


### Examine Differences by Month

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  mutate(
    month = lubridate::floor_date(Sensing_Date, "month")
  ) %>%
  count(month, sort = TRUE)
```


### Examine Differences by Sensor


```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  count(Location_ID, sort = TRUE)
```


### Inspect Largest Differences

```{r}
comparison_2022 %>%
  filter(count_difference != 0) %>%
  arrange(desc(abs(count_difference))) %>%
  select(
    Location_ID,
    Sensing_Date,
    HourDay,
    PedestrianCount,
    pedestrian_count,
    count_difference
  ) %>%
  head(20)
```


### Investigate Location 6


```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  summarise(
    n = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    min_difference = min(count_difference),
    max_difference = max(count_difference)
  )
```


### Location 6 Differences by Month

```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  mutate(
    month = lubridate::floor_date(Sensing_Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    different_records = n(),
    mean_difference = mean(count_difference),
    median_difference = median(count_difference),
    .groups = "drop"
  )
```

### Direction of Differences

```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  count(
    direction = case_when(
      count_difference > 0 ~ "Snapshot higher",
      count_difference < 0 ~ "Snapshot lower",
      TRUE ~ "Same"
    )
  )
```


### Validate Direction Totals


```{r}
comparison_2022 %>%
  filter(Location_ID == 6, count_difference != 0) %>%
  arrange(desc(abs(count_difference))) %>%
  select(
    Location_ID,
    Sensing_Date,
    HourDay,
    PedestrianCount,
    pedestrian_count,
    count_difference
  ) %>%
  head(20)
```

```{r}
snapshot_2022_full <- data.table::fread(
  file.path(snapshot_dir, "2022_Pedestrian_Data.csv")
)

snapshot_2022_full %>%
  mutate(
    direction_total = Direction_1 + Direction_2,
    difference = PedestrianCount - direction_total
  ) %>%
  summarise(
    rows = n(),
    exact_match = sum(difference == 0, na.rm = TRUE),
    different = sum(difference != 0, na.rm = TRUE),
    max_difference = max(abs(difference), na.rm = TRUE)
  )
```

```{r}
snapshot_2022_full %>%
  filter(Location_ID == 6) %>%
  summarise(
    mean_pedestrian_count = mean(PedestrianCount, na.rm = TRUE),
    mean_direction_total = mean(Direction_1 + Direction_2, na.rm = TRUE),
    mean_difference = mean(
      PedestrianCount - (Direction_1 + Direction_2),
      na.rm = TRUE
    )
  )
```

## Validate the 2024 Snapshot
### Load 2024 Snapshot

```{r}
snapshot_2024 <- data.table::fread(
  file.path(snapshot_dir, "2024_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)
```

### Compare with Existing 2024 Data

```{r}
comparison_2024 <- snapshot_2024 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    pedestrian_full %>%
      filter(lubridate::year(sensing_date) == 2024) %>%
      select(
        location_id,
        sensing_date,
        hourday,
        pedestrian_count
      ),
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    difference = PedestrianCount - pedestrian_count
  )

comparison_2024 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(difference != 0, na.rm = TRUE),
    mean_difference = mean(difference, na.rm = TRUE),
    max_absolute_difference = max(abs(difference), na.rm = TRUE)
  )
```

## Validate the 2025 Snapshot

### Load 2025 Snapshot

```{r}

snapshot_2025 <- data.table::fread(
  file.path(snapshot_dir, "2025_Pedestrian_Data.csv"),
  select = c(
    "Location_ID",
    "Sensing_Date",
    "HourDay",
    "PedestrianCount"
  )
)
```


```{r}
comparison_2025 <- snapshot_2025 %>%
  as_tibble() %>%
  mutate(
    Sensing_Date = as.Date(Sensing_Date)
  ) %>%
  inner_join(
    pedestrian_full %>%
      filter(lubridate::year(sensing_date) == 2025) %>%
      select(
        location_id,
        sensing_date,
        hourday,
        pedestrian_count
      ),
    by = c(
      "Location_ID" = "location_id",
      "Sensing_Date" = "sensing_date",
      "HourDay" = "hourday"
    )
  ) %>%
  mutate(
    difference = PedestrianCount - pedestrian_count
  )


comparison_2025 %>%
  summarise(
    matched_rows = n(),
    different_rows = sum(difference != 0, na.rm = TRUE),
    mean_difference = mean(difference, na.rm = TRUE),
    max_absolute_difference = max(abs(difference), na.rm = TRUE)
  )
```


## Standardisation and Integration

### Standardise Snapshot Data

```{r}
snapshot_standardised <- purrr::map_dfr(
  snapshot_files,
  function(f) {
    
    
    x <- data.table::fread(file = f) %>%
      as_tibble()
    
    x %>%
      transmute(
        location_id = as.numeric(Location_ID),
        sensing_date = as.Date(Sensing_Date),
        hourday = as.numeric(HourDay),
        pedestrian_count = as.numeric(PedestrianCount),
        hour_repaired = FALSE,
        source = paste0("city_snapshot_", substr(basename(f), 1, 4))
      )
  }
) %>%
  mutate(
    year = lubridate::year(sensing_date),
    month = lubridate::month(sensing_date),
    month_name = lubridate::month(sensing_date, label = TRUE),
    day_name = lubridate::wday(sensing_date, label = TRUE),
    day_of_week = lubridate::wday(sensing_date, week_start = 1),
    is_weekend = day_of_week >= 6,
    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  )
```


### Check Standardised Snapshot Data


```{r}
snapshot_standardised %>%
  count(year)
```

```{r}
snapshot_standardised %>%
  summarise(
    rows = n(),
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    sensors = n_distinct(location_id),
    duplicate_groups = n() -
      n_distinct(location_id, sensing_date, hourday)
  )
```

## Replace Existing Data

### Replace 2022–2025 with Validated Snapshots

```{r}


## Rebuild final pedestrian dataset

### Keep historical data before 2022

historical_data <- pedestrian_full %>%
  filter(!year %in% c(2022, 2023, 2024, 2025)) %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count,
    source,
    any_of("hour_repaired")
  )

# If the historical dataset does not contain hour_repaired,
# create it as FALSE.
if (!"hour_repaired" %in% names(historical_data)) {
  historical_data$hour_repaired <- FALSE
}


### Prepare validated 2022–2025 snapshots

snapshot_data <- snapshot_standardised %>%
  select(
    location_id,
    sensing_date,
    hourday,
    pedestrian_count,
    source
  ) %>%
  mutate(
    hour_repaired = FALSE
  )


### Combine historical and validated data

pedestrian_full_updated <- bind_rows(
  historical_data,
  snapshot_data
) %>%
  mutate(
    year = lubridate::year(sensing_date),

    month = lubridate::month(sensing_date),

    month_name = lubridate::month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),

    day_name = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    ),

    day_of_week = lubridate::wday(
      sensing_date,
      week_start = 1
    ),

    is_weekend = day_of_week >= 6,

    season = case_when(
      month %in% c(12, 1, 2) ~ "Summer",
      month %in% c(3, 4, 5) ~ "Autumn",
      month %in% c(6, 7, 8) ~ "Winter",
      month %in% c(9, 10, 11) ~ "Spring"
    )
  ) %>%
  arrange(
    sensing_date,
    hourday,
    location_id
)

```

### Recreate Derived Variables

```{r}
pedestrian_full_updated %>%
  group_by(year) %>%
  summarise(
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    observations = n(),
    sensors = n_distinct(location_id),
    .groups = "drop"
  )
```

## Final Validation

### Check Yearly Coverage

```{r}
final_check <- pedestrian_full_updated %>%
  summarise(
    total_rows = n(),
    unique_sensor_hour_records =
      n_distinct(location_id, sensing_date, hourday),
    duplicate_rows =
      n() - n_distinct(location_id, sensing_date, hourday),
    missing_counts =
      sum(is.na(pedestrian_count)),
    first_date = min(sensing_date),
    last_date = max(sensing_date)
  )

final_check
```

```{r}
pedestrian_full_updated %>%
  count(year)
```

```{r}
pedestrian_full <- pedestrian_full_updated
```


## Daily and Weekly Pedestrian Rhythm

### Hourly Pedestrian Activity Over Time

```{r}
library(plotly)
```

```{r}
hourly_by_sensor_year <- pedestrian_full %>%
  filter(
    !is.na(hourday),
    hourday >= 0,
    hourday <= 23
  ) %>%
  group_by(year, location_id, hourday) %>%
  summarise(
    sensor_hour_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

hourly_by_year <- hourly_by_sensor_year %>%
  group_by(year, hourday) %>%
  summarise(
    avg_pedestrians = mean(sensor_hour_avg, na.rm = TRUE),
    .groups = "drop"
  )

hourly_index <- hourly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()

hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
hourly_summary <- hourly_index %>%
  group_by(hourday) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
daily_interactive_data <- hourly_index %>%
  mutate(
    year_label = case_when(
      year == 2009 ~ "2009 (partial)",
      year == 2026 ~ "2026 (partial)",
      TRUE ~ as.character(year)
    )
  )
```

### PLOT 1 — Typical Daily Pedestrian Rhythm

```{r}

daily_interactive <- plot_ly(
  daily_interactive_data,
  x = ~hourday,
  y = ~activity_index,
  color = ~year_label,
  type = "scatter",
  mode = "lines",
  text = ~paste0(
    "Year: ", year_label,
    "<br>Hour: ", hourday, ":00",
    "<br>Activity index: ", round(activity_index, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Daily Pedestrian Rhythm by Year",
    xaxis = list(
      title = "Hour of day",
      tickmode = "linear",
      tick0 = 0,
      dtick = 1
    ),
    yaxis = list(
      title = "Activity index (100 = yearly average)"
    ),
    hovermode = "closest",
    legend = list(
      title = list(text = "Year"),
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    margin = list(r = 150)
  )

daily_interactive
```

## Weekly Rhythm Across the Years, 2009–2026

```{r}

pedestrian_full <- pedestrian_full %>%
  mutate(
    day_name = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1
    )
  )
```

### Calculating Weekly Patterns

```{r}
weekly_by_sensor_year <- pedestrian_full %>%
  group_by(year, location_id, day_name) %>%
  summarise(
    sensor_day_avg = mean(pedestrian_count, na.rm = TRUE),
    .groups = "drop"
  )

weekly_by_year <- weekly_by_sensor_year %>%
  group_by(year, day_name) %>%
  summarise(
    avg_pedestrians = mean(sensor_day_avg, na.rm = TRUE),
    .groups = "drop"
  )

weekly_index <- weekly_by_year %>%
  group_by(year) %>%
  mutate(
    yearly_mean = mean(avg_pedestrians, na.rm = TRUE),
    activity_index = avg_pedestrians / yearly_mean * 100
  ) %>%
  ungroup()

weekly_summary <- weekly_index %>%
  group_by(day_name) %>%
  summarise(
    median_index = median(activity_index, na.rm = TRUE),
    q25 = quantile(activity_index, 0.25, na.rm = TRUE),
    q75 = quantile(activity_index, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
weekly_interactive_data <- weekly_index %>%
  mutate(
    year_label = case_when(
      year == 2009 ~ "2009 (partial)",
      year == 2026 ~ "2026 (partial)",
      TRUE ~ as.character(year)
    )
  )
```

### PLOT 2 — Typical Weekly Pedestrian Rhythm

```{r}
weekly_interactive <- plot_ly(
  weekly_interactive_data,
  x = ~day_name,
  y = ~activity_index,
  color = ~year_label,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    "Year: ", year_label,
    "<br>Day: ", day_name,
    "<br>Activity index: ", round(activity_index, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Weekly Pedestrian Rhythm by Year",
    xaxis = list(
      title = "Day of week"
    ),
    yaxis = list(
      title = "Activity index (100 = yearly average)"
    ),
    hovermode = "closest",
    legend = list(
      title = list(text = "Year"),
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    margin = list(r = 150)
  )

weekly_interactive
```


### ### Daily pedestrian activity for dip analysis

```{r}

decrease_sensor_daily <- pedestrian_full %>%
  filter(
    !is.na(pedestrian_count)
  ) %>%
  group_by(
    location_id,
    sensing_date
  ) %>%
  summarise(
    daily_count = sum(
      pedestrian_count,
      na.rm = TRUE
    ),
    hours_observed = n_distinct(hourday),
    .groups = "drop"
  ) %>%
  filter(
    hours_observed >= 20
  )
```

### Calculate expected activity for each sensor and weekday

```{r}


decrease_sensor_index <- decrease_sensor_daily %>%
  arrange(
    location_id,
    sensing_date
  ) %>%
  mutate(
    day_of_week = lubridate::wday(
      sensing_date,
      week_start = 1
    )
  ) %>%
  group_by(
    location_id,
    day_of_week
  ) %>%
  mutate(
    expected_count = slider::slide_dbl(
      daily_count,
      ~ median(
        .x[-5],
        na.rm = TRUE
      ),
      .before = 4,
      .after = 4,
      .complete = TRUE
    )
  ) %>%
  ungroup() %>%
  mutate(
    decrease_percent =
      100 *
      (daily_count - expected_count) /
      expected_count
  ) %>%
  filter(
    is.finite(decrease_percent)
  )
```


### Summarise pedestrian decreases across sensors


```{r}


decrease_day_summary <- decrease_sensor_index %>%
  group_by(
    sensing_date
  ) %>%
  summarise(
    median_decrease = median(
      decrease_percent,
      na.rm = TRUE
    ),

    sensors_reporting = n_distinct(
      location_id
    ),

    sensors_below_20 = sum(
      decrease_percent <= -20,
      na.rm = TRUE
    ),

    proportion_below_20 = mean(
      decrease_percent <= -20,
      na.rm = TRUE
    ),

    median_daily_count = median(
      daily_count,
      na.rm = TRUE
    ),

    .groups = "drop"
  )
```

### Identify significant pedestrian decreases

```{r}

decrease_candidates <- decrease_day_summary %>%
  filter(
    sensors_reporting >= 10,
    median_decrease <= -20,
    proportion_below_20 >= 0.30
  ) %>%
  arrange(
    median_decrease
  )

decrease_candidates %>%
  slice_head(n = 30)
```

### Focus on non-COVID significant decreases

```{r}

decrease_non_covid <- decrease_candidates %>%
  filter(
    lubridate::year(sensing_date) < 2020 |
      lubridate::year(sensing_date) > 2022
  )
```


```{r}
decrease_non_covid %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20
  ) %>%
  slice_head(n = 30)
```

```{r}
decrease_non_covid %>%
  mutate(
    year = lubridate::year(sensing_date)
  ) %>%
  count(year)
```



```{r}
### Top non-COVID pedestrian decreases

top_decreases <- decrease_non_covid %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20
  ) %>%
  slice_head(n = 20) %>%
  mutate(
    day = lubridate::wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    )
  )

top_decreases
```



```{r}
top_decreases %>%
  mutate(
    month = lubridate::month(
      sensing_date,
      label = TRUE,
      abbr = FALSE
    )
  ) %>%
  count(month, sort = TRUE)
```

```{r}
top_decreases %>%
  mutate(
    month_day = format(
      sensing_date,
      "%d %B"
    )
  ) %>%
  count(month_day, sort = TRUE)
```


```{r}
top_decreases %>%
  select(
    sensing_date,
    median_decrease,
    sensors_reporting,
    proportion_below_20,
    day
  ) %>%
  arrange(sensing_date)
```


```{r}
### Victorian public holidays, 2009–2026

easter_sunday <- function(y) {
  a <- y %% 19
  b <- y %/% 100
  c <- y %% 100
  d <- b %/% 4
  e <- b %% 4
  f <- (b + 8) %/% 25
  g <- (b - f + 1) %/% 3
  h <- (19 * a + b - d - g + 15) %% 30
  i <- c %/% 4
  k <- c %% 4
  l <- (32 + 2 * e + 2 * i - h - k) %% 7
  m <- (a + 11 * h + 22 * l) %/% 451

  month <- (h + l - 7 * m + 114) %/% 31
  day <- ((h + l - 7 * m + 114) %% 31) + 1

  as.Date(
    sprintf("%04d-%02d-%02d", y, month, day)
  )
}
```


### Load Victorian Government Public Holiday Data

```{r}

holiday_calendar_official <- readr::read_csv(
  "../Data_pedestrians/raw2/important_dates_august_2026.csv",
  show_col_types = FALSE
) %>%
  filter(dateType == "PUBLIC_HOLIDAY") %>%
  transmute(
    holiday_date = lubridate::dmy(important_date),
    holiday = name,
    publisher,
    source
  ) %>%
  filter(
    holiday_date >= as.Date("2019-01-01"),
    holiday_date <= as.Date("2026-09-06")
  ) %>%
  distinct(holiday_date, .keep_all = TRUE) %>%
  arrange(holiday_date)

holiday_calendar_official %>%
  count(lubridate::year(holiday_date), name = "holidays")
```



```{r}

decrease_holiday_test <- decrease_non_covid %>%
  filter(
    lubridate::year(sensing_date) %in%
      c(2019, 2023, 2024, 2025, 2026)
  ) %>%
  left_join(
    holiday_calendar_official %>%
      select(holiday_date, holiday),
    by = c("sensing_date" = "holiday_date")
  ) %>%
  mutate(
    holiday_status = if_else(
      is.na(holiday),
      "Normal day",
      "Public holiday"
    )
  )

decrease_holiday_test %>%
  count(holiday_status) %>%
  mutate(
    percentage = round(n / sum(n) * 100, 1)
  )
```

### Compare public holidays with all eligible days

```{r}

eligible_days <- decrease_day_summary %>%
  filter(
    sensors_reporting >= 10,
    lubridate::year(sensing_date) %in%
      c(2019, 2023, 2024, 2025, 2026)
  ) %>%
  distinct(sensing_date) %>%
  mutate(
    holiday_status = if_else(
      sensing_date %in% holiday_calendar_official$holiday_date,
      "Public holiday",
      "Normal day"
    )
  )

baseline_holiday <- eligible_days %>%
  count(holiday_status) %>%
  mutate(
    percentage = round(n / sum(n) * 100, 1)
  )

baseline_holiday
```


```{r}
### Public holiday enrichment using official government data

holiday_enrichment <- tibble::tibble(
  baseline_holiday_percentage =
    mean(eligible_days$holiday_status == "Public holiday") * 100,

  significant_dip_holiday_percentage =
    mean(decrease_holiday_test$holiday_status == "Public holiday") * 100
) %>%
  mutate(
    enrichment =
      significant_dip_holiday_percentage /
      baseline_holiday_percentage
  )

holiday_enrichment
```


```{r}
### Identify Public Holidays Associated with Significant Decreases

public_holiday_dips <- decrease_holiday_test %>%
  filter(holiday_status == "Public holiday") %>%
  select(
    sensing_date,
    holiday,
    median_decrease,
    sensors_reporting
  ) %>%
  arrange(median_decrease)

knitr::kable(
  public_holiday_dips,
  digits = 1,
  caption = "Significant Pedestrian Decreases on Victorian Public Holidays"
)
```


```{r}
### Interactive Significant Pedestrian Decreases Across Years

holiday_timeline_data <- decrease_holiday_test %>%
  mutate(
    holiday_name = if_else(
      is.na(holiday),
      "Not a public holiday",
      holiday
    ),

    hover_text = paste0(
      "Date: ", format(sensing_date, "%d %B %Y"),
      "<br>Holiday: ", holiday_name,
      "<br>Median decrease: ", round(median_decrease, 1), "%",
      "<br>Sensors reporting: ", sensors_reporting
    )
  ) %>%
  arrange(sensing_date)

holiday_timeline <- plotly::plot_ly(
  holiday_timeline_data,
  x = ~sensing_date,
  y = ~median_decrease,
  color = ~holiday_status,
  colors = c(
    "Normal day" = "#64748B",
    "Public holiday" = "#E87932"
  ),
  type = "scatter",
  mode = "markers",
  marker = list(
    size = 11,
    opacity = 0.85
  ),
  text = ~hover_text,
  hoverinfo = "text"
) %>%
  layout(
    title = "Significant Pedestrian Decreases and Public Holidays",
    xaxis = list(
      title = "Date"
    ),
    yaxis = list(
      title = "Median change from expected activity (%)"
    ),
    hovermode = "closest",
    legend = list(
      title = list(text = "Day type")
    )
  )

holiday_timeline
```


## Weather and Pedestrian Activity Analysis

```{r}

weather_daily <- readr::read_csv(
  "../Data_pedestrians/raw2/melbourne_weather_daily.csv",
  skip = 3,
  show_col_types = FALSE
) %>%
  transmute(
    sensing_date = as.Date(time),

    max_temperature_c = `temperature_2m_max (°C)`,

    min_temperature_c = `temperature_2m_min (°C)`,

    max_wind_kmh = `wind_speed_10m_max (km/h)`,

    rainfall_mm = `precipitation_sum (mm)`
  ) %>%
  filter(
    sensing_date >= as.Date("2009-05-01"),
    sensing_date <= as.Date("2026-09-06")
  )

weather_daily %>%
  summarise(
    total_days = n(),
    first_date = min(sensing_date),
    last_date = max(sensing_date),
    missing_rainfall = sum(is.na(rainfall_mm)),
    duplicate_dates = n() - n_distinct(sensing_date)
  )
```

### Combine Weather with Pedestrian Decreases

```{r}

weather_dip_daily <- decrease_day_summary %>%
  filter(
    sensors_reporting >= 10,
    lubridate::year(sensing_date) %in%
      c(2019, 2023, 2024, 2025, 2026)
  ) %>%
  left_join(
    weather_daily,
    by = "sensing_date"
  ) %>%
  mutate(
    significant_dip =
      median_decrease <= -20 &
      proportion_below_20 >= 0.30,

    public_holiday =
      sensing_date %in%
      holiday_calendar_official$holiday_date
  )

weather_dip_daily %>%
  summarise(
    total_days = n(),
    significant_dip_days = sum(significant_dip),
    public_holiday_days = sum(public_holiday),
    holiday_dip_days = sum(significant_dip & public_holiday),
    missing_weather = sum(is.na(rainfall_mm)),
    duplicate_dates = n() - n_distinct(sensing_date)
  )
```


```{r}
weather_dip_daily %>%
  summarise(
    missing_weather = sum(is.na(rainfall_mm)),
    duplicate_dates = n() - n_distinct(sensing_date)
  )
```



```{r}
### Compare Weather on Significant Dip Days and Other Days

weather_comparison <- weather_dip_daily %>%
  filter(!public_holiday) %>%
  group_by(significant_dip) %>%
  summarise(
    days = n(),

    median_rainfall_mm =
      round(median(rainfall_mm, na.rm = TRUE), 1),

    heavy_rain_pct =
      round(mean(rainfall_mm >= 10, na.rm = TRUE) * 100, 1),

    median_max_temp_c =
      round(median(max_temperature_c, na.rm = TRUE), 1),

    median_min_temp_c =
      round(median(min_temperature_c, na.rm = TRUE), 1),

    median_max_wind_kmh =
      round(median(max_wind_kmh, na.rm = TRUE), 1),

    .groups = "drop"
  )

print(weather_comparison, width = Inf)
```


```{r}
weather_comparison %>%
  select(
    significant_dip,
    days,
    median_min_temp_c,
    median_max_wind_kmh
  )
```


### Compare Pedestrian Dip Frequency on Heavy Rainfall Days

```{r}
rain_dip_frequency <- weather_dip_daily %>%
  filter(!public_holiday) %>%
  mutate(
    rain_category = if_else(
      rainfall_mm >= 10,
      "Heavy rainfall (10 mm or more)",
      "Less than 10 mm rainfall"
    )
  ) %>%
  group_by(rain_category) %>%
  summarise(
    total_days = n(),

    significant_dip_days = sum(significant_dip),

    dip_percentage = round(
      mean(significant_dip) * 100,
      1
    ),

    .groups = "drop"
  )

print(rain_dip_frequency, width = Inf)
```

### Significant Pedestrian Decreases by Rainfall and Season

```{r}
rain_season_comparison <- weather_dip_daily %>%
  filter(!public_holiday) %>%
  mutate(
    season = case_when(
      lubridate::month(sensing_date) %in% c(12, 1, 2) ~ "Summer",
      lubridate::month(sensing_date) %in% c(3, 4, 5) ~ "Autumn",
      lubridate::month(sensing_date) %in% c(6, 7, 8) ~ "Winter",
      TRUE ~ "Spring"
    ),
    rain_category = if_else(
      rainfall_mm >= 10,
      "Heavy rainfall",
      "Less than 10 mm"
    )
  ) %>%
  group_by(season, rain_category) %>%
  summarise(
    total_days = n(),
    significant_dip_days = sum(significant_dip),
    dip_percentage = round(
      mean(significant_dip) * 100, 1
    ),
    .groups = "drop"
  )

print(rain_season_comparison, n = Inf)
```


### Interactive Rainfall and Pedestrian Decreases by Season

```{r}
### Interactive Rainfall and Pedestrian Decreases by Season

rain_season_plot_data <- rain_season_comparison %>%
  mutate(
    season = factor(
      season,
      levels = c("Summer", "Autumn", "Winter", "Spring")
    ),
    hover_text = paste0(
      "Season: ", season,
      "<br>Rainfall: ", rain_category,
      "<br>Total days: ", total_days,
      "<br>Significant dip days: ", significant_dip_days,
      "<br>Dip rate: ", dip_percentage, "%"
    )
  )

# Create the interactive graph
rain_season_plot <- plotly::plot_ly(
  data = rain_season_plot_data,
  x = ~season,
  y = ~dip_percentage,
  color = ~rain_category,
  type = "bar",
  text = ~hover_text,
hoverinfo = "text",
textposition = "none"
)

# Add graph titles and formatting
rain_season_plot <- plotly::layout(
  rain_season_plot,
  title = "Significant Pedestrian Decreases by Rainfall and Season",
  xaxis = list(title = "Season"),
  yaxis = list(
    title = "Significant decrease rate (%)",
    range = c(0, 40)
  ),
  barmode = "group",
  legend = list(
    title = list(text = "Rainfall category")
  )
)

rain_season_plot
```

### Significant Pedestrian Decreases by Rainfall Amount

```{r}
rainfall_levels <- weather_dip_daily %>%
  filter(!public_holiday) %>%
  mutate(
    rainfall_group = case_when(
      rainfall_mm == 0 ~ "No rain",
      rainfall_mm < 2 ~ "0–2 mm",
      rainfall_mm < 5 ~ "2–5 mm",
      rainfall_mm < 10 ~ "5–10 mm",
      rainfall_mm < 20 ~ "10–20 mm",
      TRUE ~ "20 mm or more"
    ),

    rainfall_group = factor(
      rainfall_group,
      levels = c(
        "No rain",
        "0–2 mm",
        "2–5 mm",
        "5–10 mm",
        "10–20 mm",
        "20 mm or more"
      )
    )
  ) %>%
  group_by(rainfall_group) %>%
  summarise(
    total_days = n(),
    significant_dip_days = sum(significant_dip),

    dip_percentage = round(
      mean(significant_dip) * 100,
      1
    ),

    .groups = "drop"
  )

print(rainfall_levels, n = Inf)
```

### Interactive Rainfall Intensity and Pedestrian Decreases

```{r}
rainfall_levels_plot <- rainfall_levels %>%
  mutate(
    hover_text = paste0(
      "Rainfall: ", rainfall_group,
      "<br>Total days: ", total_days,
      "<br>Significant dip days: ", significant_dip_days,
      "<br>Dip rate: ", dip_percentage, "%"
    )
  )

rainfall_interactive <- plotly::plot_ly(
  data = rainfall_levels_plot,
  x = ~rainfall_group,
  y = ~dip_percentage,
  type = "bar",
  marker = list(color = "#478FAD"),
  hovertext = ~hover_text,
  hoverinfo = "text"
)

rainfall_interactive <- plotly::layout(
  rainfall_interactive,
  title = "Significant Pedestrian Decreases by Daily Rainfall",
  xaxis = list(
    title = "Daily rainfall (mm)"
  ),
  yaxis = list(
    title = "Significant decrease rate (%)",
    range = c(0, 55)
  ),
  showlegend = FALSE
)

rainfall_interactive
```


### Combined Weather Comparison

```{r}
weather_overview <- weather_comparison %>%
  mutate(
    day_type = if_else(
      significant_dip,
      "Significant dip days",
      "Other days"
    )
  ) %>%
  select(
    day_type,
    median_rainfall_mm,
    heavy_rain_pct,
    median_max_temp_c,
    median_max_wind_kmh
  ) %>%
  tidyr::pivot_longer(
    cols = -day_type,
    names_to = "weather_factor",
    values_to = "value"
  ) %>%
  mutate(
    weather_factor = dplyr::recode(
      weather_factor,
      median_rainfall_mm = "Median rainfall (mm)",
      heavy_rain_pct = "Heavy rainfall days (%)",
      median_max_temp_c = "Maximum temperature (°C)",
      median_min_temp_c = "Minimum temperature (°C)",
      median_max_wind_kmh = "Maximum wind speed (km/h)"
    )
  )

weather_overview_plot <- ggplot(
  weather_overview,
  aes(
    x = day_type,
    y = value,
    fill = day_type
  )
) +
  geom_col(width = 0.6) +
  facet_wrap(
    ~weather_factor,
    scales = "free_y",
    ncol = 2
  ) +
  labs(
    title = "Weather Conditions and Pedestrian Decreases",
    subtitle = "Non-public holidays, 2019 and 2023–2026",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

plotly::ggplotly(weather_overview_plot)
```


### Combined Weather Factors and Pedestrian Decreases

```{r}
weather_model_data <- weather_dip_daily %>%
  filter(!public_holiday) %>%
  mutate(
    heavy_rain = rainfall_mm >= 10,
    season = case_when(
      lubridate::month(sensing_date) %in% c(12, 1, 2) ~ "Summer",
      lubridate::month(sensing_date) %in% c(3, 4, 5) ~ "Autumn",
      lubridate::month(sensing_date) %in% c(6, 7, 8) ~ "Winter",
      TRUE ~ "Spring"
    )
  )

weather_model <- glm(
  significant_dip ~ heavy_rain +
    max_temperature_c +
    max_wind_kmh +
    factor(season),
  data = weather_model_data,
  family = binomial
)

model_summary <- summary(weather_model)$coefficients

data.frame(
  factor = rownames(model_summary),
  odds_ratio = round(exp(model_summary[, "Estimate"]), 2),
  p_value = signif(model_summary[, "Pr(>|z|)"], 3),
  row.names = NULL
)
```

```{r}
### Check Confidence Intervals for Weather Model

weather_model_ci <- exp(
  cbind(
    Odds_Ratio = coef(weather_model),
    confint.default(weather_model)
  )
)

round(weather_model_ci[-1, ], 2)
```




```{r}
#| label: prepare-yearly-weather-holiday-data

# Combine weather, pedestrian changes and holiday names

weather_holiday_yearly <- weather_dip_daily %>%
  left_join(
    holiday_calendar_official %>%
      group_by(holiday_date) %>%
      summarise(
        holiday = paste(unique(holiday), collapse = " / "),
        .groups = "drop"
      ),
    by = c("sensing_date" = "holiday_date")
  ) %>%
  mutate(
    year_label = factor(
      as.character(lubridate::year(sensing_date)),
      levels = c("2019", "2023", "2024", "2025", "2026")
    ),

    # Align all years by month and day for comparison
    plot_date = as.Date(
      paste0("2024-", format(sensing_date, "%m-%d"))
    ),

    holiday_name = coalesce(
      holiday,
      "Not a public holiday"
    ),

    hover_text = paste0(
      "Date: ", format(sensing_date, "%d %B %Y"),
      "<br>Holiday: ", holiday_name,
      "<br>Median pedestrian change: ",
      round(median_decrease, 1), "%",
      "<br>Rainfall: ", rainfall_mm, " mm",
      "<br>Maximum temperature: ", max_temperature_c, "°C",
      "<br>Minimum temperature: ", min_temperature_c, "°C",
      "<br>Maximum wind speed: ", max_wind_kmh, " km/h",
      "<br>Sensors reporting: ", sensors_reporting
    )
  ) %>%
  arrange(year_label, plot_date)
```



```{r}
### Smoothed Pedestrian Changes, Weather and Holidays by Year

smoothed_yearly_data <- weather_holiday_yearly %>%
  group_by(year_label) %>%
  arrange(sensing_date, .by_group = TRUE) %>%
  mutate(
    smoothed_change = slider::slide_dbl(
      median_decrease,
      ~ median(.x, na.rm = TRUE),
      .before = 3,
      .after = 3
    )
  ) %>%
  ungroup()

# Coloured line for each year
smoothed_yearly_plot <- plotly::plot_ly(
  data = smoothed_yearly_data,
  x = ~plot_date,
  y = ~smoothed_change,
  color = ~year_label,
  type = "scatter",
  mode = "lines",
  line = list(width = 2),
  text = ~paste0(
    hover_text,
    "<br>7-day median change: ",
    round(smoothed_change, 1), "%"
  ),
  hoverinfo = "text"
)

# Actual significant decreases on public holidays
smoothed_yearly_plot <- smoothed_yearly_plot %>%
  plotly::add_markers(
    data = smoothed_yearly_data %>%
      filter(significant_dip, public_holiday),
    x = ~plot_date,
    y = ~median_decrease,
    text = ~hover_text,
    hoverinfo = "text",
    name = "Public holiday dip",
    marker = list(color = "#E87932", size = 9),
    inherit = FALSE
  )

# Actual significant decreases on other dates
smoothed_yearly_plot <- smoothed_yearly_plot %>%
  plotly::add_markers(
    data = smoothed_yearly_data %>%
      filter(significant_dip, !public_holiday),
    x = ~plot_date,
    y = ~median_decrease,
    text = ~hover_text,
    hoverinfo = "text",
    name = "Other significant dip",
    marker = list(color = "#3288BD", size = 8),
    inherit = FALSE
  )

# Month labels from January to December
month_ticks <- seq.Date(
  as.Date("2024-01-01"),
  as.Date("2024-12-01"),
  by = "month"
)

smoothed_yearly_plot <- plotly::layout(
  smoothed_yearly_plot,
  title = list(
  text = paste0(
    "Pedestrian Changes, Weather and Public Holidays by Year",
    "<br><sup>Lines: 7-day rolling median | ",
    "Dots: actual significant pedestrian decreases</sup>"
  )
),
  xaxis = list(
    title = "Month of year",
    tickvals = month_ticks,
    ticktext = format(month_ticks, "%b"),
    range = c("2024-01-01", "2024-12-31")
  ),
  yaxis = list(
    title = "Change from expected pedestrian activity (%)"
  ),
  hovermode = "closest",
  legend = list(
    title = list(text = "Year / Dip type")
  )
)

smoothed_yearly_plot
```




```{r}
sensor_locations <- readr::read_csv(
  "../Data_pedestrians/raw/data_pedestrian/2025_Pedestrian_Data.csv",
  col_select = c(
    Location_ID,
    SensorVariable,
    Latitude,
    Longitude
  ),
  show_col_types = FALSE
) %>%
  distinct(Location_ID, .keep_all = TRUE)
```


### Load Sensor Location Information

```{r}
# Create one location record per sensor

sensor_lookup <- sensor_locations %>%
  group_by(Location_ID) %>%
  summarise(
    sensor_name = first(SensorVariable),
    latitude = first(Latitude),
    longitude = first(Longitude),
    .groups = "drop"
  ) %>%
  rename(location_id = Location_ID)


# Identify sensors affected on significant decrease dates

sensor_dip_summary <- decrease_sensor_index %>%
  semi_join(
    weather_dip_daily %>%
      filter(
        significant_dip,
        lubridate::year(sensing_date) %in% 2023:2025
      ) %>%
      distinct(sensing_date),
    by = "sensing_date"
  ) %>%
  filter(!is.na(decrease_percent)) %>%
  group_by(location_id) %>%
  summarise(
    days_reporting = n_distinct(sensing_date),

    sensor_dip_days = sum(decrease_percent <= -20),

    affected_pct = round(
      100 * sensor_dip_days / days_reporting, 1
    ),

    median_change_pct = round(
      median(decrease_percent), 1
    ),

    .groups = "drop"
  ) %>%
  filter(days_reporting >= 10) %>%
  left_join(sensor_lookup, by = "location_id") %>%
  arrange(desc(affected_pct), desc(sensor_dip_days))


# Display the 15 most frequently affected sensors

sensor_dip_summary %>%
  select(
    location_id,
    sensor_name,
    days_reporting,
    sensor_dip_days,
    affected_pct,
    median_change_pct
  ) %>%
  slice_head(n = 15) %>%
  print(n = 15, width = Inf)


# Check whether any sensors could not be matched

sum(is.na(sensor_dip_summary$sensor_name))
```


```{r}
sensor_overall_dips <- decrease_sensor_index %>%
  filter(
    sensing_date >= as.Date("2023-01-01"),
    sensing_date <= as.Date("2025-12-31"),
    !is.na(decrease_percent),
    is.finite(decrease_percent)
  ) %>%
  group_by(location_id) %>%
  summarise(
    days_reporting = n_distinct(sensing_date),

    dip_days = sum(decrease_percent <= -20),

    dip_rate_pct = round(
      100 * mean(decrease_percent <= -20), 1
    ),

    median_dip_pct = round(
      median(
        decrease_percent[decrease_percent <= -20],
        na.rm = TRUE
      ), 1
    ),

    .groups = "drop"
  ) %>%
  filter(days_reporting >= 180) %>%
  left_join(sensor_lookup, by = "location_id") %>%
  arrange(desc(dip_rate_pct))

print(
  sensor_overall_dips %>%
    select(
      location_id,
      sensor_name,
      days_reporting,
      dip_days,
      dip_rate_pct,
      median_dip_pct
    ) %>%
    slice_head(n = 15),
  n = 15,
  width = Inf
)
```

```{r}

# Correct paths for the 2023 and 2024 files

older_sensor_files <- file.path(
  "..",
  "Data_pedestrians",
  "raw",
  "data_pedestrian",
  paste0(2023:2024, "_Pedestrian_Data.csv")
)

file.exists(older_sensor_files)
```


```{r}
sensor_7_details <- purrr::map_dfr(
  older_sensor_files,
  function(f) {
    readr::read_csv(
      f,
      col_select = c(
        Location_ID,
        SensorVariable,
        Latitude,
        Longitude
      ),
      show_col_types = FALSE
    ) %>%
      filter(Location_ID == 7) %>%
      distinct() %>%
      mutate(source_file = basename(f))
  }
)

print(sensor_7_details, n = Inf, width = Inf)
```

```{r}
# Add sensor 7 to our sensor lookup

sensor_7_lookup <- sensor_7_details %>%
  distinct(Location_ID, .keep_all = TRUE) %>%
  transmute(
    location_id = Location_ID,
    sensor_name = SensorVariable,
    latitude = Latitude,
    longitude = Longitude
  )

sensor_lookup <- sensor_lookup %>%
  filter(location_id != 7) %>%
  bind_rows(sensor_7_lookup)

# Update the existing results with complete sensor information

sensor_overall_dips <- sensor_overall_dips %>%
  select(-any_of(c("sensor_name", "latitude", "longitude"))) %>%
  left_join(sensor_lookup, by = "location_id")

# Verify the correction

sensor_overall_dips %>%
  filter(location_id == 7) %>%
  select(
    location_id, sensor_name,
    dip_days, dip_rate_pct,
    latitude, longitude
  )

# Check for any remaining missing coordinates

sum(is.na(sensor_overall_dips$latitude) |
    is.na(sensor_overall_dips$longitude))
```

```{r}
sensor_overall_dips %>%
  filter(
    is.na(latitude) | is.na(longitude)
  ) %>%
  select(
    location_id,
    sensor_name,
    days_reporting,
    dip_days,
    dip_rate_pct,
    latitude,
    longitude
  ) %>%
  print(n = Inf, width = Inf)
```


```{r}
# Check older pedestrian datasets for missing sensor locations

missing_sensor_ids <- c(122, 65, 26, 55)

sensor_history_files <- file.path(
  "..",
  "Data_pedestrians",
  "raw",
  "data_pedestrian",
  paste0(2022:2024, "_Pedestrian_Data.csv")
)

missing_sensor_details <- purrr::map_dfr(
  sensor_history_files,
  function(f) {
    readr::read_csv(
      f,
      col_select = c(
        Location_ID,
        SensorVariable,
        Latitude,
        Longitude
      ),
      show_col_types = FALSE
    ) %>%
      filter(Location_ID %in% missing_sensor_ids) %>%
      distinct() %>%
      mutate(source_file = basename(f))
  }
)

print(missing_sensor_details, n = Inf, width = Inf)
```


```{r}
recovered_sensor_locations <- missing_sensor_details %>%
  transmute(
    location_id = Location_ID,
    sensor_name = SensorVariable,
    latitude = Latitude,
    longitude = Longitude
  ) %>%
  distinct()

print(recovered_sensor_locations, n = Inf)
```


```{r}
sensor_lookup_final <- bind_rows(
  sensor_locations %>%
    transmute(
      location_id = Location_ID,
      sensor_name = SensorVariable,
      latitude = Latitude,
      longitude = Longitude
    ),
  recovered_sensor_locations
) %>%
  distinct(location_id, .keep_all = TRUE)
```


```{r}
sensor_lookup_final %>%
  filter(location_id %in% c(122, 65, 26, 55)) %>%
  print(n = Inf)
```


```{r}
sensor_dip_map_ready <- sensor_dip_summary %>%
  left_join(
    sensor_lookup_final,
    by = "location_id"
  )
```


```{r}
# Combine existing sensor locations with the 4 recovered sensors
sensor_lookup_final <- bind_rows(
  sensor_lookup,
  recovered_sensor_locations
) %>%
  distinct(location_id, .keep_all = TRUE)

# Join locations to the overall sensor dip analysis
sensor_dip_map_ready <- sensor_overall_dips %>%
  select(
    -any_of(c("sensor_name", "latitude", "longitude"))
  ) %>%
  left_join(
    sensor_lookup_final,
    by = "location_id"
  )
```


```{r}
sensor_dip_map_ready %>%
  filter(is.na(latitude) | is.na(longitude)) %>%
  print(n = Inf)
```


```{r}

sensor_dip_map_ready %>%
  select(
    location_id,
    sensor_name,
    days_reporting,
    dip_days,
    dip_rate_pct,
    latitude,
    longitude
  ) %>%
  arrange(desc(dip_rate_pct)) %>%
  print(n = 15, width = Inf)
```


```{r}
#| label: sensor-dip-map

# Install leaflet once if necessary
if (!requireNamespace("leaflet", quietly = TRUE)) {
  install.packages("leaflet")
}

# Create colour scale
sensor_palette <- leaflet::colorNumeric(
  palette = "YlOrRd",
  domain = sensor_dip_map_ready$dip_rate_pct
)

# Interactive map
sensor_dip_map <- leaflet::leaflet(
  data = sensor_dip_map_ready
) %>%
  leaflet::addTiles() %>%
  leaflet::addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = ~sensor_palette(dip_rate_pct),
    fillColor = ~sensor_palette(dip_rate_pct),
    fillOpacity = 0.8,
    weight = 1,
    label = ~paste0(
      sensor_name,
      " | Dip rate: ", dip_rate_pct, "%"
    ),
    popup = ~paste0(
      "<b>Sensor: </b>", sensor_name,
      "<br><b>Location ID: </b>", location_id,
      "<br><b>Reporting days: </b>", days_reporting,
      "<br><b>Dip days: </b>", dip_days,
      "<br><b>Dip rate: </b>", dip_rate_pct, "%",
      "<br><b>Median dip: </b>", median_dip_pct, "%"
    )
  ) %>%
  leaflet::addLegend(
    position = "bottomleft",
    pal = sensor_palette,
    values = ~dip_rate_pct,
    title = "Dip rate (%)",
    opacity = 0.8
  )

sensor_dip_map

```


```{r}
#| label: sensor-weather-holiday-comparison

# Combine sensor decreases with weather and holiday information

sensor_condition_data <- decrease_sensor_index %>%
  filter(
    sensing_date >= as.Date("2023-01-01"),
    sensing_date <= as.Date("2025-12-31"),
    is.finite(decrease_percent),
    location_id %in% sensor_dip_map_ready$location_id
  ) %>%
  inner_join(
    weather_dip_daily %>%
      select(sensing_date, rainfall_mm, public_holiday),
    by = "sensing_date"
  ) %>%
  mutate(
    condition = case_when(
      public_holiday ~ "Public holiday",
      rainfall_mm >= 10 ~ "Heavy rainfall",
      TRUE ~ "Other days"
    ),
    sensor_dip = decrease_percent <= -20
  )

# Calculate how frequently each sensor decreased

sensor_condition_summary <- sensor_condition_data %>%
  group_by(location_id, condition) %>%
  summarise(
    days_reporting = n_distinct(sensing_date),
    dip_days = sum(sensor_dip),
    dip_rate_pct = round(
      100 * mean(sensor_dip), 1
    ),
    .groups = "drop"
  ) %>%
  left_join(
    sensor_lookup_final %>%
      select(location_id, sensor_name),
    by = "location_id"
  ) %>%
  filter(days_reporting >= 15)

# Show the five highest dip rates for each condition

sensor_condition_summary %>%
  group_by(condition) %>%
  slice_max(
    dip_rate_pct,
    n = 5,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(
    condition,
    sensor_name,
    days_reporting,
    dip_days,
    dip_rate_pct
  ) %>%
  print(n = Inf, width = Inf)

```


```{r}
sensor_condition_summary %>%
  filter(condition == "Public holiday") %>%
  arrange(desc(dip_rate_pct)) %>%
  select(
    sensor_name,
    days_reporting,
    dip_days,
    dip_rate_pct
  ) %>%
  slice_head(n = 10) %>%
  print(n = Inf, width = Inf)
```


```{r}
# Compare the same sensors across different conditions

comparison_sensors <- c(
  "NewQ_T", "FedCycle_T", "WatCit_T",
  "Pel147_T", "UM1_T", "UM2_T",
  "UM3_T", "KenMac_T",
  "FedPed_T", "BirFed1120_T"
)

sensor_final_comparison <- sensor_condition_summary %>%
  filter(sensor_name %in% comparison_sensors) %>%
  mutate(
    condition = recode(
      condition,
      "Heavy rainfall" = "Rain",
      "Public holiday" = "Holiday",
      "Other days" = "Other"
    )
  ) %>%
  select(
    location_id,
    sensor_name,
    condition,
    days_reporting,
    dip_rate_pct
  ) %>%
  tidyr::pivot_wider(
    names_from = condition,
    values_from = c(dip_rate_pct, days_reporting),
    names_glue = "{.value}_{condition}"
  ) %>%
  arrange(sensor_name)

print(sensor_final_comparison, n = Inf, width = Inf)
```


```{r}
sensor_final_comparison %>%
  transmute(
    Sensor = sensor_name,
    Rain = dip_rate_pct_Rain,
    Holiday = dip_rate_pct_Holiday,
    Other = dip_rate_pct_Other
  ) %>%
  print(n = Inf, width = Inf)
```

```{r}
# Identify specific public holidays associated with sensor dips

sensor_holiday_breakdown <- decrease_sensor_index %>%
  filter(
    sensing_date >= as.Date("2023-01-01"),
    sensing_date <= as.Date("2025-12-31"),
    is.finite(decrease_percent),
    location_id %in% sensor_final_comparison$location_id
  ) %>%
  inner_join(
    holiday_calendar_official %>%
      group_by(holiday_date) %>%
      summarise(
        holiday = paste(unique(holiday), collapse = " / "),
        .groups = "drop"
      ),
    by = c("sensing_date" = "holiday_date")
  ) %>%
  group_by(location_id, holiday) %>%
  summarise(
    holiday_days = n_distinct(sensing_date),
    dip_days = sum(decrease_percent <= -20),
    dip_rate_pct = round(
      100 * dip_days / holiday_days, 1
    ),
    .groups = "drop"
  ) %>%
  left_join(
    sensor_lookup_final %>%
      select(location_id, sensor_name),
    by = "location_id"
  ) %>%
  arrange(sensor_name, desc(dip_days))

# Look at four interesting sensors

sensor_holiday_breakdown %>%
  filter(
    sensor_name %in% c(
      "Pel147_T", "UM1_T",
      "NewQ_T", "WatCit_T"
    )
  ) %>%
  select(
    sensor_name,
    holiday,
    holiday_days,
    dip_days,
    dip_rate_pct
  ) %>%
  print(n = Inf, width = Inf)
```


```{r}
sensor_holiday_breakdown %>%
  dplyr::filter(sensor_name == "Pel147_T") %>%
  dplyr::select(
    holiday,
    holiday_days,
    dip_days,
    dip_rate_pct
  ) %>%
  as.data.frame() %>%
  print(row.names = FALSE)
```



